---
title: "Calling peaks"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this tutorial, we demonstrate how to call peaks on a single-cell ATAC-seq
dataset using [MACS2](https://github.com/macs3-project/MACS).

To use the peak calling functionality in Signac you will first need to install
MACS2. This can be done using [pip](https://pypi.org/project/MACS2/) or
[conda](https://anaconda.org/bioconda/macs2), or by building the package
from [source](https://github.com/macs3-project/MACS).

In this demonstration we use scATAC-seq data for human PBMCs. See our
[vignette](pbmc_vignette.Rmd) for the code used to generate this object,
and links to the raw data. First, load the required packages and the
pre-computed Seurat object:

```{r message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)

pbmc <- readRDS("../vignette_data/pbmc.rds")
DimPlot(pbmc)
```

Peak calling can be performed using the `CallPeaks()` function, and can either
be done separately for different groups of cells, or performed using data from
all the cells. To call peaks on each annotated cell type, we can use the 
`group.by` argument:

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(
  object = pbmc,
  group.by = "predicted.id",
  macs2.path = "/home/stuartt/miniconda3/envs/signac/bin/macs2"
)
```

The results are returned as a `GRanges` object, with an additional metadata column
listing the cell types that each peak was identified in:

```{r echo=FALSE}
knitr::kable(head(as.data.frame(peaks)))
```

To quantify counts in each peak, you can use the `FeatureMatrix()` function.

We can visualize the cell-type-specific MACS2 peak calls alongside the 10x Cellranger
peak calls (currently being used in the `pbmc` object) with the `CoveragePlot()`
function. Here the Cellranger peaks are shown in grey and the MACS2 peaks in red:

```{r fig.height=6, fig.width=10}
CoveragePlot(
  object = pbmc,
  region = "CD8A",
  ranges = peaks,
  ranges.title = "MACS2"
)
```
<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
