---
title: "Data structures and object interaction"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

The Signac package is an extension of [Seurat](https://satijalab.org/seurat/)
designed for the analysis of single-cell chromatin assays. This includes assays
that generate signal mapped to genomic coordinates, such as scATAC-seq,
scCUT&Tag, scNTT-seq, and other methods.

As the analysis of these single-cell chromatin datasets presents some unique 
challenges in comparison to the analysis of scRNA-seq data, we have created an
extended `Assay5` class to store the additional information needed, including:

* Genomic ranges associated with the features (eg, peaks or genomic bins)
* Gene annotations
* Genome information 
* TF motifs 
* Genome-wide signal in a disk-based format (fragment files) 
* TF footprinting data 
* Tn5 insertion bias data 
* Linked genomic regions 

A major advantage of the Signac design is its interoperability with existing
functions in the Seurat package, and other packages that are able to use the
Seurat object. This enables straightforward analysis of multimodal single-cell
data through the addition of different assays to the Seurat object.

Here we outline the design of each class defined in the Signac package, and
demonstrate methods that can be run on each class.

## Differences between Signac v1 and v2

We first introduced Signac with the [version 1 release](https://doi.org/10.1038/s41592-021-01282-5)
in 2020. Recently we have released Signac v2, with some major updates to the
package. The main changes include:

* **Updating the `ChromatinAssay` to use the Seurat `Assay5` class:** this
update enables storing data in the object using different matrix types, including
as a [BPCells](https://github.com/bnprks/BPCells) matrix. BPCells provides
efficient on-disk matrix storage and computation methods, and enables scaling
analyses to millions of cells with a small memory requirement.  
* **More flexible object classes**: we now provide options to use Signac assays
without needing each feature to correspond to a single region of the genome.
This is useful for storing different types of feature, including our recently-released
universal feature set for the human genome, [REMO](https://github.com/stuart-lab/REMO.v1.GRCh38)  
* **More flexible storage of genomic links**: you can now store multiple sets of
genomic links within the same assay. We have also moved to storing this information
using the well-established `GInteractions` format from the [InteractionSet](https://www.bioconductor.org/packages/InteractionSet/)
package.  
* **New peak calling methods**: Signac now uses `macs3` in the `CallPeaks()`. We
have also enabled peak parallelization of peak calls across groups of cells, and
include an option to use the `hmmratac` peak calling method in `macs3`.  
* **New data visualization methods**: we added the ability to plot GWAS and QTL
data alongside single-cell chromatin data in the `CoveragePlot()` function

For a full changelog please see the package [release notes](news.html).

To update old `ChromatinAssay` assays you can use the `as.GRangesAssay()` 
function in Signac v2.

## The `ChromatinAssay5` Class

The `ChromatinAssay5` class extends the standard Seurat `Assay5` class and adds
several additional slots for data useful for the analysis of single-cell
chromatin datasets. The class includes all the slots present in a standard 
Seurat object with the following additional information:
    
* `motifs`: A `Motif` object
* `fragments`: A list of `Fragment` objects
* `annotation`: A [`GRanges`](https://www.rdocumentation.org/packages/GenomicRanges/versions/1.24.1/topics/GRanges-class)
object containing gene annotations
* `bias`: A vector containing Tn5 integration bias information (the frequency of
Tn5 integration at different hexamers)
* `region.aggregation`: A named list of `RegionAggregation` objects containing 
signal per cell aggregated across a set of genomic regions (for example, motif
sites)
* `links`: A named list of [`GInteractions`](https://www.bioconductor.org/packages/release/bioc/html/InteractionSet.html)
objects describing linked genomic regions, such as co-accessible sites or
enhancer-gene regulatory relationships.

## The `GRangesAssay` Class

The `GRangesAssay` is an extended `ChromatinAssay5` object class that adds a
requirement that the features stored in the assay correspond to genomic ranges.
It stores the genomic ranges associated with features in the assay as a
[`GRanges`](https://www.rdocumentation.org/packages/GenomicRanges/versions/1.24.1/topics/GRanges-class)
object. The `GRangesAssay` enables some additional analyses that require that
the assay features correspond to genomic regions, such as peak-gene linkage or
motif enrichment.

Here we will focus on demonstrating use of the `GRangesAssay`, but all functions
that run on a `GRangesAssay` object can also be run on a `ChromatinAssay5` object.
There are some cases where a `ChromatinAssay5` object would be the preferred way
to store your data. For example, you may wish to create an assay that does not
store genomic peak intervals (perhaps quantifying [REMO features](https://github.com/stuart-lab/REMO.v1.GRCh38)
or transcription factor activities), but still want the ability to create
genome coverage plots.

### Constructing the `GRangesAssay`

```{r message=FALSE, warning=FALSE}
library(Seurat)
library(Signac)
```

A `GRangesAssay` object can be constructed using the
`CreateGRangesAssay()` function.

```{r}
# get some data to use in the following examples
counts <- LayerData(atac_small, layer = "counts")

# create a standalone GRangesAssay object
gr_assay <- CreateGRangesAssay(counts = counts)
```

To create a Seurat object that contains a `GRangesAssay` rather than a
standard `Assay`, we can initialize the object using the `GRangesAssay` rather
than a count matrix.

```{r}
# create a Seurat object containing a GRangesAssay
object <- CreateSeuratObject(counts = gr_assay)
```

### Adding a `GRangesAssay` to a `Seurat` object

To add a new `GRangesAssay` object to an existing Seurat object, we can use
the standard assignment operation used for adding standard `Assay` objects and 
other data types to the Seurat object.

```{r}
object[["peaks"]] <- CreateGRangesAssay(counts = counts)
```

### Getting and setting `GRangesAssay` data

We can get/set data for the `GRangesAssay` in much the same way we do for a
standard `Assay` object: using the `LayerData` function
defined in `Seurat`. For example:

```{r}
## Getting

# access the data slot, found in standard Assays and GRangesAssays
data <- LayerData(atac_small, layer = "data")

# access the bias slot, unique to the GRangesAssay
bias <- LayerData(atac_small, layer = "bias")

## Setting

# set the data slot
LayerData(atac_small, layer = "data") <- data
```

We also have a variety of convenience functions defined for getting/setting 
data in specific slots. This includes the `Fragments()`, `Motifs()`, `Links()`,
`Annotation()`, `RegionAggr()`, and `Bias()` functions. For example, to get or
set gene annotation data we can use the `Annotation()` getter and `Annotation<-`
setter functions:

```{r message=FALSE, warning=FALSE}
# first get some gene annotations
library(AnnotationHub)
ah <- AnnotationHub()

ensdb_v98 <- ah[["AH75011"]]

# convert EnsDb to GRanges
gene.ranges <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

# set gene annotations
Annotation(atac_small) <- gene.ranges

# get gene annotation information
Annotation(atac_small)
```

The `Fragments()`, `Motifs()`, and `Links()` functions are demonstrated in other sections
below.

### Other `GRangesAssay` methods

As the `GRangesAssay` object uses the Bioconductor  [`GRanges`](https://www.rdocumentation.org/packages/GenomicRanges/versions/1.24.1/topics/GRanges-class)
class to store information, we can also call standard Bioconductor functions defined in the 
`IRanges`, `GenomicRanges`, and `Seqinfo` packages on the `GRangesAssay`
object (or a Seurat object with a `GRangesAssay` as the default assay).

The following methods use the genomic ranges stored in a `GRangesAssay` object.

```{r message=FALSE, warning=FALSE}
# extract the genomic ranges associated with each feature in the data matrix
granges(atac_small)

# find the nearest range
nearest(atac_small, subject = Annotation(atac_small))

# distance to the nearest range
distanceToNearest(atac_small, subject = Annotation(atac_small))

# find overlaps with another set of genomic ranges
findOverlaps(atac_small, subject = Annotation(atac_small))
```

Many other methods are defined, see the documentation for `nearest-methods`,
`findOverlaps-methods`, `inter-range-methods`, and `coverage` in Signac for a
full list.

For a full list of methods for the `GRangesAssay` class run:

```{r}
methods(class = "GRangesAssay")
```

### Subsetting a `GRangesAssay`

We can use the standard `subset()` function or the `[` operator to subset Seurat
object containing `GRangesAssay`s. This works the same way as for standard
`Assay` objects.

```{r}
# subset using the subset() function
# this is meant for interactive use
subset.obj <- subset(atac_small, subset = nCount_peaks > 10)

# subset using the [ extract operator
# this can be used programmatically
subset.obj <- atac_small[, atac_small$nCount_peaks > 10]
```

### Converting between `Assay` and `GRangesAssay` or `ChromatinAssay5`

To convert from a `GRangesAssay` to a standard `Assay` use the `as()` function

```{r}
# convert a ChromatinAssay to an Assay
assay <- as(object = atac_small[["peaks"]], Class = "Assay5")
assay
```

To convert from a standard `Assay` to a `GRangesAssay` we use the 
`as.GRangesAssay()` function. This takes a standard assay object, as well as
information to fill the additional slots in the `GRangesAssay` class.

```{r}
# convert an Assay to a ChromatinAssay
gr_assay <- as.GRangesAssay(assay)
gr_assay
```

Similarly, we can use `as.ChromatinAssay5()` to convert to a `ChromatinAssay5`
object:

```{r}
chr_assay <- as.ChromatinAssay5(assay)
chr_assay
```

## The `Fragment2` Class

The `Fragment2` class is designed for storing and interacting with a
[fragment file](https://support.10xgenomics.com/single-cell-atac/software/pipelines/latest/output/fragments)
commonly used for single-cell chromatin data. It contains:
* The path to an indexed fragment file on disk  
* The path to the fragment file index 
* An MD5 hash for the fragment file and the fragment file index
* A vector of cell names contained in the fragment file. Importantly,
this is a named vector where the elements of the vector are the cell names as 
they appear in the fragment file, and the name of each element is the cell
name as it appears in the assay object storing the `Fragment2`
object. This allows a mapping of cell names on disk to cell names in R, and 
avoids the need to alter fragment files on disk. This path can also be a remote
file accessible by `http` or `ftp`.
* A vector of sequence names contained in the fragment file. This is a named 
vector of sequence levels (eg, chromosome name) where each element is the
sequence name as it appears in the fragment file, and then name of each element
is the corresponding sequence name as stored in the assay. This enables renaming
sequence names in your assay object without needing to change your fragment file.

### Constructing the `Fragment2` class

A `Fragment` object can be constructed using the `CreateFragmentObject()`
function. 

```{r}
frag.path <- system.file(
  "extdata", "fragments_header.tsv.gz", package = "Signac"
)
fragments <- CreateFragmentObject(
  path = frag.path,
  cells = colnames(atac_small),
  validate.fragments = TRUE
)
```

The `validate.fragments` parameter controls whether the file is inspected to
check whether the expected cell names are present. This can help avoid assigning
the wrong fragment file to the object. If you're sure that the file is correct,
you can set this value to `FALSE` to skip this step and save some time. This
check is typically only run once when the `Fragment2` object is created, and is 
not normally run on existing `Fragment2` files.

### Inspecting the fragment file

To extract the first few lines of a fragment file on-disk, we can use the
`head()` method defined for `Fragment` objects. This is useful for quickly
checking the chromosome naming style in our fragment file, or checking how the 
cell barcodes are named:

```{r}
head(fragments)
```

Some fragment files also contain header lines with additional information. This
can be viewed using the `header()` function:

```{r}
header(fragments)
```
### Adding a `Fragment2` object to the `GRangesAssay`

A `GRangesAssay` object can contain a list of `Fragment2` objects. This avoids
the need to merge fragment files on disk and simplifies processes of merging
or integrating different Seurat objects containing `GRangesAssay`s. To add 
a new `Fragment2` object to a `GRangesAssay`, or a Seurat object containing a 
`GRangesAssay`, we can use the `Fragments<-` assignment function. This will
do a few things:

1. Re-compute the MD5 hash for the fragment file and index and verify that it
matches the hash computed when the `Fragment` object was created.
2. Check that none of the cells contained in the `Fragment` object being added
are already contained in another `Fragment` object stored in the
`GRangesAssay`. All fragments from a cell must be present in only one fragment
file.
3. Check that the same fragment file is not already stored in the assay.  
4. Append the `Fragment` object to the list of `Fragment` objects stored in the
`GRangesAssay`.

```{r}
Fragments(atac_small) <- fragments
```

The `show()` method for `Fragment`-class objects prints the number of cells that
the `Fragment` object contains data for.

```{r}
fragments
```

Alternatively, we can initialize the `GRangesAssay` with a `Fragment` object
in a couple of ways. We can either pass a vector of `Fragment` objects to the
`fragments` parameter in `CreateGRangesAssay()`, or pass the path to a
single fragment file. If we pass the path to a fragment file we assume that the
file contains fragments for all cells in the `GRangesAssay` and that the cell
names are the same in the fragment file on disk and in the `GRangesAssay`.
For example:

```{r}
chrom_assay <- CreateGRangesAssay(
  counts = counts,
  fragments = frag.path
)
object <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "peaks"
)
```

This will create a Seurat object containing a `GRangesAssay`, with a single
`Fragment2` object.

### Removing a `Fragment2` object from the `GRangesAssay`

All the `Fragment2` objects associated with a `GRangesAssay` can be removed
by assigning `NULL` using the `Fragment<-` assignment function. For example:

```{r}
Fragments(chrom_assay) <- NULL
Fragments(chrom_assay)
```

### Changing the fragment file path in an existing `Fragment2` object

The path to the fragment file can be updated using the `UpdatePath()` function.
This can be useful if you move the fragment file to a new directory, or if you
copy a stored Seurat object containing a `GRangesAssay` to a different server.

```{r message=FALSE}
fragments <- UpdatePath(fragments, new.path = frag.path)
```

To change the path to fragment files in an object, you will need to remove
the fragment objects, update the paths, and then add the fragment objects
back to the object. For example:

```{r}
frags <- Fragments(object) # get list of fragment objects
Fragments(object) <- NULL # remove fragment information from assay

# create a vector with all the new paths, in the correct order for your
# list of fragment objects
# In this case we only have 1
new.paths <- list(frag.path)
for (i in seq_along(frags)) {
  frags[[i]] <- UpdatePath(frags[[i]], new.path = new.paths[[i]]) # update path
}

Fragments(object) <- frags # assign updated list back to the object
Fragments(object)
```

### Using remote fragment files

Fragment files hosted on remote servers accessible via http or ftp can also be
added to the `GRangesAssay` in the same way as for locally-hosted fragment 
files. This can enable the exploration of large single-cell datasets without
the need for downloading large files. For example, we can create a Fragment
object using a file hosted on the 10x Genomics website:

```{r}
fragments <- CreateFragmentObject(
  path = "http://cf.10xgenomics.com/samples/cell-atac/1.1.0/atac_v1_pbmc_10k/atac_v1_pbmc_10k_fragments.tsv.gz"
)
fragments
```

When files are hosted remotely, the checks described in the section above (MD5
hash and expected cells) are not performed.

### Getting and setting `Fragment2` data

To access the cell names stored in a `Fragment2` object, we can use the `Cells()`
function. **Importantly**, this returns the cell names as they appear in the 
`GRangesAssay`, rather than as they appear in the fragment file itself.

```{r}
fragments <- CreateFragmentObject(
  path = frag.path,
  cells = colnames(atac_small),
  validate.fragments = TRUE
)

cells <- Cells(fragments)
head(cells)
```

Similarly, we can set the cell name information in a `Fragment2` object using the
`Cells<-` assignment function. This will set the named vector of cells stored in
the `Fragment2` object. Here we must supply a named vector.

```{r}
names(cells) <- cells
Cells(fragments) <- cells
```

To extract any of the data stored in a `Fragment` object we can also use the
`GetFragmentData()` function. For example, we can find the path to the fragment
file on disk:

```{r}
GetFragmentData(object = fragments, slot = "file.path")
```

For a full list of methods for the `Fragment2` class run:

```{r}
methods(class = "Fragment2")
```

## The `Motif` Class

The `Motif` class stores information needed for DNA sequence motif analysis, and
has the following slots:

* `data`: a sparse feature by motif matrix, where entries are 1 if the feature
contains the motif, and 0 otherwise
* `pwm`: A named list of position weight or position frequency matrices
* `motif.names`: a list of motif IDs and their common names
* `positions`: A `GRangesList` object containing the exact positions of each
motif
* `meta.data`: Additional information about the motifs 

Many of these slots are optional and do not need to be filled, but are only
required when running certain functions. For example, the `positions` slot
will be needed if running TF footprinting.

### Constructing the `Motif` class

A `Motif` object can be constructed using the `CreateMotifObject()` function.
Much of the data needed for constructing a `Motif` object can be generated using
functions from the [TFBSTools](https://www.bioconductor.org/packages/release/bioc/html/TFBSTools.html)
and [motifmatchr](https://www.bioconductor.org/packages/release/bioc/html/motifmatchr.html)
packages. Position frequency matrices for motifs can be loaded using the
[JASPAR](http://jaspar.genereg.net/) packages on
[Bioconductor](https://bioconductor.org/packages/release/data/annotation/html/JASPAR2020.html)
or the [chromVARmotifs](https://github.com/GreenleafLab/chromVARmotifs) package.
For example:

```{r include=FALSE}
if (!requireNamespace("JASPAR2020", quietly = TRUE)) {
  BiocManager::install("JASPAR2020")
}
if (!requireNamespace("TFBSTools", quietly = TRUE)) {
  BiocManager::install("TFBSTools")
}
if (!requireNamespace("motifmatchr", quietly = TRUE)) {
  BiocManager::install("motifmatchr")
}
if (!requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)) {
  BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
}
```

```{r message=FALSE, warning=FALSE}
library(JASPAR2020)
library(TFBSTools)
library(motifmatchr)

# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(species = 9606) # 9606 is the species code for human
)

# Scan the DNA sequence of each peak for the presence of each motif
motif.matrix <- CreateMotifMatrix(
  features = granges(atac_small),
  pwm = pfm,
  genome = "hg38"
)

# Create a new Mofif object to store the results
motif <- CreateMotifObject(
  data = motif.matrix,
  pwm = pfm
)
```

The `show()` method for the `Motif` class prints the total number of motifs and
regions included in the object:

```{r}
motif
```

### Adding a `Motif` object to the `GRangesAssay`

We can add a `Motif` object to the `GRangesAssay`, or a Seurat object
containing a `GRangesAssay` using the `Motifs<-` assignment operator.

```{r warning=FALSE}
Motifs(atac_small) <- motif
```

### Getting and setting `Motif` data

Data stored in a `Motif` object can be accessed using the `GetMotifData()` and 
`SetMotifData()` functions.

```{r}
# extract data from the Motif object
pfm <- GetMotifData(object = motif, slot = "pwm")

# set data in the Motif object
motif <- SetMotifData(object = motif, slot = "pwm", new.data = pfm)
```

We can access the set of motifs and set of features used in the `Motif` object
using the `colnames()` and `rownames()` functions:

```{r}
# look at the motifs included in the Motif object
head(colnames(motif))
```

```{r}
# look at the features included in the Motif object
head(rownames(motif))
```

To quickly convert between motif IDs (like `MA0497.1`) and motif common names 
(like MEF2C), we can use the `ConvertMotifID()` function. For example:

```{r}
# convert ID to common name
ids <- c("MA0030.1", "MA0031.1", "MA0051.1", "MA0057.1")
names <- ConvertMotifID(object = motif, id = ids)
names
```

```{r}
# convert names to IDs
ConvertMotifID(object = motif, name = names)
```

For a full list of methods for the `Motif` class run:

```{r warning=FALSE, message=FALSE}
methods(class = "Motif")
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>

