---
title: "Frequently asked questions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is Signac?

Signac is an extension of Seurat for the analysis of single-cell chromatin data
(DNA-based single-cell assays). We have extended the Seurat object to include
information about the genome sequence and genomic coordinates of sequenced
fragments per cell, and include functions needed for the analysis of single-cell
chromatin data.

## How do I interact with the object?

Signac uses the Seurat object structure, and so all the Seurat commands can be
used when analysing data with Signac. See the Data Structures and Object
Interaction [vignette](data_structures.html) for an explanation of the classes defined in Signac
and how to use them. See the Seurat documentation for more information about the
Seurat object: https://satijalab.org/seurat/

## How do I merge objects with Signac?

See the [merge](merging.html) and [integration](integration.html) vignettes for
information on combining multiple single-cell chromatin datasets.

## How can I create a fragment file for my dataset?

The fragment file is provided in the output of popular single-cell data
data processing tools such as [chromap](https://github.com/haowenz/chromap), [cellranger-atac](https://support.10xgenomics.com/single-cell-atac/software/pipelines/latest/what-is-cell-ranger-atac),
and [cellranger-arc](https://support.10xgenomics.com/single-cell-multiome-atac-gex/software/pipelines/latest/what-is-cell-ranger-arc).

If you are using another method that does not provide a fragment file as output,
you can use the [sinto](https://github.com/timoast/sinto) package to generate
a fragment file from the BAM file. See here for more information on using Sinto
to generate a fragment file:
https://timoast.github.io/sinto/basic_usage.html#create-scatac-seq-fragments-file

## How should I decide on the number of dimensions to use?

Choosing the dimensionality is a general problem in single-cell analysis for
which there is no simple solution. There has been discussion about this for
scRNA-seq, and you can read our recommendations for scRNA-seq in the Seurat
vignettes: https://satijalab.org/seurat/v3.1/pbmc3k_tutorial.html (see
"Determine the ‘dimensionality’ of the dataset").

Here are some general tips/suggestions that might help guide you in the choice
for number of dimensions:

* the number of dimensions needed will generally scale with the size and
complexity of the dataset
* you can try varying the number of dimensions used and observing how the
resulting clusters or UMAP changes
* it is usually better to choose values that are higher rather than too low
* having a good understanding of the biology will help a lot in knowing whether
the clusters make sense, or if the dimensionality might be too high/low

## An annotation or genome sequence for my organism is not available on Bioconductor, what do I do?

If you are studying an organism that does not have a `BSgenome` genome package
or `EnsDB` annotation package available on BioConductor, you can still use your
own GTF file or FASTA files with Signac.

To create your own `BSgenome` data package, see this
[vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/BSgenome/inst/doc/BSgenomeForge.pdf).

To use a GTF file, you can import it using `rtracklayer`, for example:

```{r, eval=FALSE}
gtf <- rtracklayer::import('genes.gtf')
gene.coords <- gtf[gtf$type == 'gene']
seqlevelsStyle(gene.coords) <- 'UCSC'
gene.coords <- keepStandardChromosomes(gene.coords, pruning.mode = 'coarse')
```

## How should I cite Signac?

If you use Signac, please cite [Stuart et al., 2021](https://doi.org/10.1038/s41592-021-01282-5):

```{r, eval=FALSE}
@ARTICLE{signac,
  title     = "Single-cell chromatin state analysis with Signac",
  author    = "Stuart, Tim and Srivastava, Avi and Madad, Shaista and Lareau,
               Caleb A and Satija, Rahul",
  journal   = "Nat. Methods",
  publisher = "Nature Publishing Group",
  pages     = "1--9",
  month     =  nov,
  year      =  2021,
  url       = "https://www.nature.com/articles/s41592-021-01282-5",
  language  = "en"
}
```

Signac is an extension of [Seurat](https://satijalab.org/seurat/), and uses the
Seurat object structure, so you should consider citing the
[Seurat paper](https://doi.org/10.1016/j.cell.2019.05.031) if you have used Signac.

