---
title: "Whole body scATAC-seq analysis with REMO and BPCells"
output: html_document
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "../../vignette_data/zhang")
```

```{r, include=FALSE}
options(width = 600)
local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) {
      options$attr.output <- c(
        options$attr.output,
        sprintf('style="max-height: %s;"', options$max.height)
      )
    }
    hook_output(x, options)
  })
})
```

<details>
  <summary>**View data download code**</summary>

```{sh eval=FALSE}
aws s3 cp s3://stuartlab/zhang/fragments.tsv.gz . --request-payer
aws s3 cp s3://stuartlab/zhang/fragments.tsv.gz.tbi . --request-payer
aws s3 cp s3://stuartlab/zhang/cells.txt.gz . --request-payer
```

</details>

In this tutorial we will demonstrate how large-scale scATAC-seq datasets from a
variety of organs can be efficiently analyzed using Signac with a small memory
footprint. This vignette analzying >620,000 cells was run on a Macbook Pro
laptop with 16 Gb RAM.

First load in Signac, Seurat, and some other packages we will be using for
analyzing human data.

```{r include=FALSE}
if (!requireNamespace("REMO.v1.GRCh38", quietly = TRUE)) {
  remotes::install_github(repo = "stuart-lab/REMO.v1.GRCh38")
}

if (!requireNamespace("BPCells", quietly = TRUE)) {
  remotes::install_github("bnprks/BPCells/r")
}
```

```{r setup, message=FALSE}
library(Signac)
library(Seurat)
library(REMO.v1.GRCh38)
library(BPCells)
library(GenomicRanges)
```

## Pre-processing workflow

When pre-processing chromatin data, Signac uses the ATAC fragment file as the
main input data. This represents a full list of all unique fragments across all
single cells. It is stored on-disk for memory efficiency and scalability. More
information about the fragment file can be found on the 10x Genomics [website](https://support.10xgenomics.com/single-cell-atac/software/pipelines/latest/output/fragments) or on the [sinto website](https://timoast.github.io/sinto/basic_usage.html#create-scatac-seq-fragments-file).

For improved scalability we will use our recently-released [REMO features]()
for the human genome, and store these features on-disk using [BPCells](https://bnprks.github.io/BPCells/index.html).

Here we have pre-computed the REMO count matrix using our [`fragtk`]() Rust software
for fast and memory-efficient matrix quantification, and start by converting this
matrix file to a BPCells matrix:

<details>
  <summary>**View `fragtk` command**</summary>

```{bash, eval=FALSE}
fragtk matrix -f fragments.tsv.gz -b REMOv1_GRCh38.bed.gz -c cells.txt.gz -o remo --pic --group
```

</details>  


```{r eval=FALSE, include=TRUE}
# import count matrix
counts <- import_matrix_market_10x(mtx_dir = "remo")

# Write the matrix to a directory in BPCells format
outpath <- paste0(normalizePath(getwd()), "/../bpcells_counts")
write_matrix_dir(mat = counts, dir = outpath)
```

# Load on-disk BPCells matrix

```{r}
remo_matrix <- open_matrix_dir(dir = "../bpcells_counts")
```

```{r}
class(remo_matrix)
```

Load gene annotations for hg38

```{r message=FALSE, warning=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()

ensdb_v98 <- ah[["AH75011"]]
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

# change to UCSC style since the data was mapped to hg38
seqlevelsStyle(annotations) <- "UCSC"
```

Create Seurat object

```{r}
chrom_assay <- CreateChromatinAssay5(
  counts = remo_matrix,
  fragments = "fragments.tsv.gz",
  annotation = annotations,
  min.cells = 20
)

fetal <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "REMO",
  names.field = 2,
  names.delim = "__"
)
```

```{r}
fetal
```

Despite containing a large number of cells, we can see that the object requires
less that 1 Gb of RAM due to storing the count matrix on disk:

```{r}
format(object.size(fetal), "Gb")
```


## Quality control

We can compute quality control metrics for the scATAC-seq data using the `ATACqc` function.
These metrics include:

* Nucleosome banding pattern: The histogram of DNA fragment sizes (determined from the paired-end sequencing reads) should exhibit a strong nucleosome banding pattern corresponding to the length of DNA wrapped around a single nucleosome. We calculate this per single cell, and quantify the approximate ratio of mononucleosomal to nucleosome-free fragments (stored as `Nucleosome_signal`)

* Transcriptional start site (TSS) enrichment score: The [ENCODE project](https://www.encodeproject.org/) has defined an ATAC-seq targeting score based on the ratio of fragments centered at the TSS to fragments in TSS-flanking regions (see https://www.encodeproject.org/data-standards/terms/). Poor ATAC-seq experiments typically will have a low TSS enrichment score. Results are stored in metadata under the column name `TSS_enrichment`.

* Total number of fragments: A measure of cellular sequencing depth / complexity. Cells with very few fragments may need to be excluded due to low sequencing depth. Cells with extremely high levels may represent doublets, nuclei clumps, or other artefacts. This is stored as
`total_fragments`

```{r}
fetal <- ATACqc(fetal)
```

```{r}
DensityScatter(
  object = fetal,
  x = "total_fragments",
  y = "TSS_enrichment",
  log_x = TRUE,
  log_y = TRUE,
  quantiles = TRUE
)
```

```{r}
fetal <- subset(fetal, TSS_enrichment > 1)
```

```{r}
fetal
```

## Feature selection

```{r}
fetal <- FitMeanVar(fetal)
```

## Normalization and linear dimensional reduction

```{r}
fetal <- NormalizeData(fetal)
fetal <- RunSVD(fetal, pca = TRUE)
```

## Non-linear dimension reduction

Now that the cells are embedded in a low-dimensional space we project the cells
into a two dimensional space for visualization using UMAP, and color the cells
by their original tissue annotation.

```{r message=FALSE, warning=FALSE}
fetal <- RunUMAP(object = fetal, dims = 1:30)

# extract original tissue annotations
fetal$tissue <- unlist(lapply(strsplit(as.character(fetal$orig.ident), "_SM-"), `[[`, 1))
DimPlot(fetal, group.by = "tissue", label = TRUE, raster = FALSE, repel = TRUE, pt.size = 0.1) + NoLegend()
```

## Plotting genomic regions

```{r message=FALSE, warning=FALSE, out.width='90%', fig.height=6}
CoveragePlot(
  object = fetal,
  group.by = "tissue",
  region = "CD8A",
  ranges = REMO.v1.GRCh38,
  ranges.group.by = "REMO",
  extend.upstream = 1000,
  extend.downstream = 1000
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qs2::qs_save(object = fetal, file = "fetal.qs2")
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
