---
title: "Large-scale scATAC-seq analysis with REMO and BPCells"
output: html_document
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "../../vignette_data/zhang")
```

```{r, include=FALSE}
options(width = 600)
local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) {
      options$attr.output <- c(
        options$attr.output,
        sprintf('style="max-height: %s;"', options$max.height)
      )
    }
    hook_output(x, options)
  })
})
```

<details>
  <summary>**View data download code**</summary>

```{sh eval=FALSE}
aws s3 cp s3://stuartlab/zhang/fragments.tsv.gz . --request-payer
aws s3 cp s3://stuartlab/zhang/fragments.tsv.gz.tbi . --request-payer
aws s3 cp s3://stuartlab/zhang/cells.txt.gz . --request-payer
```

</details>

In this tutorial we will demonstrate how large-scale scATAC-seq datasets from a
variety of organs can be efficiently analyzed using Signac with a small memory
footprint. This vignette analzying >620,000 cells was run on a Macbook Pro
laptop with 16 Gb RAM.

First load in Signac, Seurat, and some other packages we will be using for
analyzing human data.

```{r include=FALSE}
if (!requireNamespace("REMO.v1.GRCh38", quietly = TRUE)) {
  remotes::install_github(repo = "stuart-lab/REMO.v1.GRCh38")
}

if (!requireNamespace("BPCells", quietly = TRUE)) {
  remotes::install_github("bnprks/BPCells/r")
}

if (!requireNamespace("DropletUtils", quietly = TRUE)) {
  BiocManager::install("DropletUtils")
}
```

```{r setup, message=FALSE}
library(Signac)
library(Seurat)
library(GenomicRanges)
library(ggplot2)
```

## Pre-processing workflow

When pre-processing chromatin data, Signac uses the ATAC fragment file as the
main input data. This represents a full list of all unique fragments across all
single cells. It is stored on-disk for memory efficiency and scalability. More
information about the fragment file can be found on the 10x Genomics [website](https://support.10xgenomics.com/single-cell-atac/software/pipelines/latest/output/fragments) or on the [sinto website](https://timoast.github.io/sinto/basic_usage.html#create-scatac-seq-fragments-file).

For improved scalability we will use our recently-released [REMO features](https://github.com/stuart-lab/REMO.v1.GRCh38)
for the human genome, and store these features on-disk using [BPCells](https://bnprks.github.io/BPCells/index.html).

You can install REMO and BPCells from GitHub:

```{r eval=FALSE}
remotes::install_github(repo = "stuart-lab/REMO.v1.GRCh38")
```

```{r eval=FALSE}
remotes::install_github("bnprks/BPCells/r")
```

```{r}
library(REMO.v1.GRCh38)
library(BPCells)
```

We start by identifying a set of cell barcodes to include in the analysis by running
the `ATACqc()` function on the fragment file:

```{r message=FALSE, warning=FALSE}
# load gene annotations for hg38
library(AnnotationHub)
ah <- AnnotationHub()

ensdb_v98 <- ah[["AH75011"]]
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)
seqlevelsStyle(annotations) <- "UCSC"

qc <- ATACqc(object = "fragments.tsv.gz", annotations = annotations)
```

```{r}
# calculate knee plot using DropletUtils
br <- DropletUtils::barcodeRanks(matrix(qc$total_fragments, nrow = 1))
df <- as.data.frame(br)
knee <- S4Vectors::metadata(br)$knee

ggplot(df, aes(rank, total)) +
  geom_line() +
  geom_hline(yintercept = knee, linetype = "dashed") +
  annotate("text", x = 1, y = knee, label = paste0("knee = ", round(knee)),
           hjust = 0, vjust = -0.5) +
  scale_x_log10() + scale_y_log10() +
  xlab("Barcode rank") + ylab("Total fragments") +
  theme_bw()
```

Next we can quantify REMO features for the selected cells using the
`FeatureMatrix()` function in Signac, storing the resulting matrix on-disk using
BPCells by setting `bpcells=TRUE`:

```{r cache = TRUE}
cells.use <- rownames(qc)[qc$total_fragments > knee]
  
counts <- FeatureMatrix(
  object = "fragments.tsv.gz",
  cells = cells.use,
  features = REMO.v1.GRCh38,
  group = TRUE,
  bpcells = TRUE,
  bpcells.dir = "./bpcells_counts"
)
```

# Creating the Seurat object

Next we initialize a Seurat object to store the data, using the Signac `ChromatinAssay5` assay class:

```{r}
chrom_assay <- CreateChromatinAssay5(
  counts = counts,
  fragments = "fragments.tsv.gz",
  annotation = annotations,
  min.cells = 20
)

fetal <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "REMO",
  names.field = 2,
  names.delim = "__",
  meta.data = qc
)
```

```{r}
fetal
```

Despite containing a large number of cells, we can see that the object requires
less that 1 Gb of RAM due to storing the count matrix on disk:

```{r}
format(object.size(fetal), "Gb")
```

## Normalization and linear dimensional reduction

We proceed with standard feature selection, normalization, and PCA steps using 
the chromatin assay containing REMO features:

```{r}
fetal <- FitMeanVar(fetal)
fetal <- NormalizeData(fetal)
fetal <- RunSVD(fetal, pca = TRUE)
```

## Non-linear dimension reduction

Now that the cells are embedded in a low-dimensional space we project the cells
into a two dimensional space for visualization using UMAP, and color the cells
by their original tissue annotation.

```{r message=FALSE, warning=FALSE}
fetal <- RunUMAP(object = fetal, dims = 1:50)

# extract original tissue annotations
fetal$tissue <- unlist(lapply(strsplit(as.character(fetal$orig.ident), "_SM-"), `[[`, 1))
DimPlot(fetal, group.by = "tissue", label = TRUE, raster = FALSE, repel = TRUE, pt.size = 0.1) + NoLegend()
```

## Plotting genomic regions

We can create genomic coverage plots for any region of the genome:

```{r fig.height=12, fig.width=14, message=FALSE, warning=FALSE}
CoveragePlot(
  object = fetal,
  group.by = "tissue",
  region = "CD8A",
  ranges = REMO.v1.GRCh38,
  ranges.group.by = "REMO",
  extend.upstream = 1000,
  extend.downstream = 1000
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qs2::qs_save(object = fetal, file = "fetal.qs2")
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
