---
title: "Parallel and distributed processing"
output: html_document
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

Parallel computing is supported in Signac through the [future](https://cran.r-project.org/package=future) package, making it easy to specify different parallelization options.
Here we demonstrate parallelization of the `FeatureMatrix` function and show some benchmark results to get a sense for the amount of speedup you might expect.

The [Seurat](https://satijalab.org/seurat/) package also uses `future` for 
parallelization, and you can see the Seurat [vignette](https://satijalab.org/seurat/v3.1/future_vignette.html) for more information.

## How to enable parallelization in Signac

Parallelization can be enabled simply by importing the `future` package and setting the `plan`.

```{r}
library(future)
plan()
```

By default the plan is set to sequential processing (no parallelization). We can change this
to `multicore`, `multiprocessor`, or `multisession` to get asynchronous processing, and set the 
number of workers to change the number of cores used.

```{r, warning=FALSE, message=FALSE}
plan("multicore", workers = 10)
plan()
```

You might also need to increase the maximum memory usage:

```{r}
options(future.globals.maxSize = 50 * 1024 ^ 3) # for 50 Gb RAM
```

Note that as of `future` version [1.14.0](https://cran.r-project.org/web/packages/future/NEWS), forked processing is disabled when running in RStudio. To enable parallel computing in RStudio, you will need to select the "multisession" option.

## Benchmarking

Here we demonstrate the runtime of `FeatureMatrix` run on 90,686 peaks for 10,247 human PBMCs under different parallelization options:

<details>
  <summary>**View benchmarking code**</summary>
  
The following code was run on Ubuntu 20.04 LTS with 56 Intel Xeon Platinum 8280L CPU @ 2.70GHz

```{bash, eval = FALSE}
# download data
wget https://cf.10xgenomics.com/samples/cell-atac/2.0.0/atac_pbmc_10k_nextgem/atac_pbmc_10k_nextgem_fragments.tsv.gz
wget https://cf.10xgenomics.com/samples/cell-atac/2.0.0/atac_pbmc_10k_nextgem/atac_pbmc_10k_nextgem_fragments.tsv.gz.tbi
wget https://cf.10xgenomics.com/samples/cell-atac/2.0.0/atac_pbmc_10k_nextgem/atac_pbmc_10k_nextgem_peaks.bed
```
  
```{r, eval=FALSE}
library(Signac)

# load data
fragments <- "../vignette_data/atac_pbmc_10k_nextgem_fragments.tsv.gz"
peaks.10k <- read.table(
  file = "../vignette_data/atac_pbmc_10k_nextgem_peaks.bed",
  col.names = c("chr", "start", "end")
)
peaks <- GenomicRanges::makeGRangesFromDataFrame(peaks.10k)
cells <- readLines("../vignette_data/pbmc10k/cells.txt")

fragments <- CreateFragmentObject(path = fragments, cells = cells, validate.fragments = FALSE)

# set number of replicates
nrep <- 5
results <- data.frame()
process_n <- 2000

# run sequentially
timing.sequential <- c()
for (i in seq_len(nrep)) {
  start <- Sys.time()
  fmat <- FeatureMatrix(fragments = fragments, features = peaks, cells = cells, process_n = process_n)
  timing.sequential <- c(timing.sequential, as.numeric(Sys.time() - start, units = "secs"))
}
res <- data.frame(
  "setting" = rep("Sequential", nrep),
  "cores" = rep(1, nrep),
  "replicate" = seq_len(nrep),
  "time" = timing.sequential
)
results <- rbind(results, res)

# 4 core
library(future)
plan("multicore", workers = 4)
options(future.globals.maxSize = 100000 * 1024^2)

timing.4core <- c()
for (i in seq_len(nrep)) {
  start <- Sys.time()
  fmat <- FeatureMatrix(fragments = fragments, features = peaks, cells = cells, process_n = process_n)
  timing.4core <- c(timing.4core, as.numeric(Sys.time() - start, units = "secs"))
}
res <- data.frame(
  "setting" = rep("Parallel", nrep),
  "cores" = rep(4, nrep),
  "replicate" = seq_len(nrep),
  "time" = timing.4core
)
results <- rbind(results, res)

# 10 core
plan("multicore", workers = 10)

timing.10core <- c()
for (i in seq_len(nrep)) {
  start <- Sys.time()
  fmat <- FeatureMatrix(fragments = fragments, features = peaks, cells = cells, process_n = process_n)
  timing.10core <- c(timing.10core, as.numeric(Sys.time() - start, units = "secs"))
}
res <- data.frame(
  "setting" = rep("Parallel", nrep),
  "cores" = rep(10, nrep),
  "replicate" = seq_len(nrep),
  "time" = timing.10core
)
results <- rbind(results, res)

# save results
write.table(
  x = results,
  file = paste0("../vignette_data/pbmc10k/timings_", Sys.Date(), ".tsv"),
  quote = FALSE,
  row.names = FALSE
)
```

</details>

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
results <- read.table("../vignette_data/pbmc10k/timings_2021-09-20.tsv", header = TRUE)
results$cores <- factor(results$cores, levels = c("1", "4", "10"))

p <- ggplot(results, aes(x = cores, y = time/60, color = cores)) +
  geom_jitter(width = 1/10) +
  ylim(c(0, 13)) +
  ylab("Time (min)") +
  xlab("Number of cores") +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("FeatureMatrix runtime")
p
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
