---
title: "Installation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

To use Signac first make sure Bioconductor is installed:

```{r}
# Install bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()

# To automatically install Bioconductor dependencies
setRepositories(ind=1:2)
```

## Current release

```{r}
install.packages("Signac")
```

## Development version

Unreleased versions of Signac can be installed from the GitHub repository using
the devtools package:

```{r}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("timoast/signac", ref = "develop")
```

## Old versions

Older Signac releases can be installed from the CRAN archive using devtools:

```{r}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

# replace "0.2.5" with the version that you want to install
devtools::install_version(package = 'Signac', version = package_version('0.2.5'))
```

## Docker

We provide docker images for Signac via [dockerhub](https://hub.docker.com/r/timoast/signac).

To pull the latest image from the command line:

```sh
docker pull timoast/signac:latest
```

To use as a base image in a new Dockerfile:

```sh
FROM timoast/signac:latest
```

## Conda

Signac can also be installed using conda. Note that if you use conda, you 
should install all packages through conda rather than R itself. Make sure to 
set up the conda channels first: http://bioconda.github.io/user/install.html#set-up-channels

```{bash}
conda install -c bioconda r-signac
```

## Installing genome assembly and gene annotation packages

It can also be useful (but not essential) to install species-specific packages 
containing genome and gene annotation information from Bioconductor.

[This](https://useast.ensembl.org/info/website/archives/assembly.html) table
from Ensembl provides a mapping of genome assembly to the corresponding gene
annotation version.

### Human hg19

```{r}
BiocManager::install(c('BSgenome.Hsapiens.UCSC.hg19', 'EnsDb.Hsapiens.v75'))
```

### Human hg38

```{r}
BiocManager::install(c('BSgenome.Hsapiens.UCSC.hg38', 'EnsDb.Hsapiens.v86'))
```

### Mouse mm10

```{r}
BiocManager::install(c('BSgenome.Mmusculus.UCSC.mm10', 'EnsDb.Mmusculus.v79'))
```
