---
title: "Joint single-cell mitochondrial DNA genotyping and DNA accessibility analysis"
author: Caleb Lareau and Tim Stuart
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    self_contained: True
---

```{r, include=FALSE}
if (!requireNamespace(package = "EnsDb.Hsapiens.v86", quietly = TRUE)) {
  BiocManager::install("EnsDb.Hsapiens.v86")
}
```

```{r init, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/charonfs/scratch/users/astar/gis/leungjy1/Projects/crc_mitoatac")
```

```{r packages, message=FALSE, warning=FALSE, echo=TRUE}
library(Signac)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(EnsDb.Hsapiens.v86)
library(ggVennDiagram)
```

Here, we take a look at two different datasets containing both DNA accessibility
measurements and mitochondrial mutation data in the same cells. One was sampled
from a patient with a colorectal cancer (CRC) tumor, and the other is from a 
polyclonal TF1 cell line. This data was produced by Lareau and Ludwig
et al. (2020), and you can read the original paper
here: https://doi.org/10.1038/s41587-020-0645-6.

Processed data files, including mitochondrial variant data for the CRC and TF1
dataset is available on Zenodo here: https://zenodo.org/record/3977808

Raw sequencing data and DNA accessibility processed files for the these datasets
are available on NCBI GEO here:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE142745

<details>
  <summary>**View data download code**</summary>

The required files can be downloaded by running the following lines in a shell:

```{bash eval=FALSE}
# ATAC data
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.filtered_peak_bc_matrix.h5
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.singlecell.csv
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.fragments.tsv.gz
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.fragments.tsv.gz.tbi

# mitochondrial allele data
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.A.txt.gz
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.C.txt.gz
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.G.txt.gz
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.T.txt.gz
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.depthTable.txt
wget https://zenodo.org/record/3977808/files/CRC_v12-mtMask_mgatk.chrM_refAllele.txt
```

</details>


# Colorectal cancer dataset

To demonstrate combined analyses of mitochondrial DNA variants and accessible
chromatin, we'll walk through a vignette analyzing cells from a primary
colorectal adenocarcinoma. The sample contains a mixture of malignant
epithelial cells and tumor infiltrating immune cells. 

## Loading the DNA accessibility data

First we load the scATAC-seq data and create a Seurat object following the 
standard workflow for scATAC-seq data.

```{r importData, message=FALSE, warning=FALSE}
# load counts and metadata from cellranger-atac
crc_dir <- ("/charonfs/scratch/users/astar/gis/leungjy1/Projects/crc_mitoatac")
counts.path <- file.path(crc_dir, 
                         "10XCR/crc/outs/filtered_peak_bc_matrix.h5")
counts <- Read10X_h5(filename = counts.path)

metadata <- read.csv(
  file = file.path(crc_dir, 
                   "10XCR/crc/outs/singlecell.csv"),
  header = TRUE,
  row.names = 1
)

# load gene annotations from Ensembl
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)

# change to UCSC style since the data was mapped to hg38 UCSC
seqlevels(annotations) <- paste0("chr", seqlevels(annotations))
genome(annotations) <- "hg38"

# create object
crc_assay <- CreateGRangesAssay(
  counts = counts,
  annotation = annotations,
  min.cells = 10,
  fragments = file.path(crc_dir,
                        "10XCR/crc/outs/fragments.tsv.gz")
)

crc <- CreateSeuratObject(
  counts = crc_assay,
  assay = "peaks",
  meta.data = metadata
)

crc[["peaks"]]
```


## Quality control

We can compute the standard quality control metrics for scATAC-seq and filter
out low-quality cells based on these metrics.

```{r message=FALSE, warning=FALSE}
# Augment QC metrics that were computed by cellranger-atac
crc$pct_reads_in_peaks <- crc$peak_region_fragments / crc$passed_filters * 100

# compute TSS enrichment score and nucleosome banding pattern
crc <- ATACqc(crc)
```

```{r fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
# visualize QC metrics for each cell
VlnPlot(crc, 
        c("TSS_enrichment", "nCount_peaks", "Nucleosome_signal", "pct_reads_in_peaks"), 
        pt.size = 0, 
        ncol = 4)
```

```{r}
# remove low-quality cells
crc <- subset(
  x = crc,
  subset = nCount_peaks > 1000 &
    nCount_peaks < 50000 &
    TSS_enrichment > 3 & 
    Nucleosome_signal < 4
)
crc
```

## Loading mgatk mitochondrial variant data

Next we can load the mitochondrial DNA variant data for these cells that was 
quantified using [mgatk](https://github.com/caleblareau/mgatk). The `ReadMGATK()`
function in Signac allows the output from `mgatk` to be read directly into R in
a convenient format for downstream analysis with Signac. `ReadMGATK()` returns a 
list containing the mitochondrial allele counts at each position & 2 dataframes 
containing the depth and reference allele. Here, we load the data
and add the counts to the Seurat object as a new assay. We also add the depth to
the metadata of the Seurat object.

```{r process_mgatk, message=FALSE, warning=FALSE, echo=TRUE}
# load mgatk output
mgatk.data <- ReadMGATK(dir = file.path(crc_dir,
                                       "mgatk/final"))

# create an assay
mgatk <- CreateAssay5Object(counts = mgatk.data$counts)

# Subset to cell present in the scATAC-seq assat
mgatk <- subset(mgatk, cells = Cells(crc))

# add assay and metadata to the seurat object
crc[["mgatk"]] <- mgatk
crc <- AddMetaData(crc, metadata = mgatk.data$depth[Cells(mgatk), ], 
                   col.name = "mtDNA_depth")
```

We can look at the mitochondrial sequencing depth for each cell, and further
subset the cells based on mitochondrial sequencing depth.

```{r fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
VlnPlot(crc, "mtDNA_depth", pt.size = 0.1) + scale_y_log10()
```

```{r}
# filter cells based on mitochondrial depth
crc <- subset(crc, mtDNA_depth >= 10)
crc
```

## Dimension reduction and clustering

Next we can run a standard dimension reduction and clustering workflow using the
scATAC-seq data to identify cell clusters.

```{r message=FALSE, warning=FALSE}
crc <- RunTFIDF(crc)
crc <- FindTopFeatures(crc, min.cutoff = 10)
crc <- RunSVD(crc)
crc <- RunUMAP(crc, reduction = "lsi", dims = 2:50)
crc <- FindNeighbors(crc, reduction = "lsi", dims = 2:50)
crc <- FindClusters(crc, resolution = 0.7, algorithm = 3)
```

```{r message=FALSE, warning=FALSE}
DimPlot(crc, label = TRUE) + NoLegend()
```

## Generate gene scores

To help interpret these clusters of cells, and assign a cell type label, we'll
estimate gene activities by summing the DNA accessibility in the gene body and
promoter region.

```{r setup_gene_activity, message=FALSE, warning=FALSE, echo=TRUE}
# compute gene accessibility
gene.activities <- GeneActivity(crc)

# add to the Seurat object as a new assay
rownames(gene.activities) <- make.unique(rownames(gene.activities))
crc[["RNA"]] <- CreateAssay5Object(counts = gene.activities)

crc <- NormalizeData(object = crc, assay = "RNA")
```

## Visualize interesting gene activity scores

We note the following markers for different cell types in the CRC dataset:

- EPCAM is a marker for epithelial cells
- TREM1 is a meyloid marker
- PTPRC = CD45 is a pan-immune cell marker
- IL1RL1 is a basophil marker
- GATA3 is a Tcell maker

```{r viz_gene_activitites, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
DefaultAssay(crc) <- "RNA"

FeaturePlot(
  object = crc,
  features = c("TREM1", "EPCAM", "PTPRC", "IL1RL1", "GATA3", "KIT"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 2
)
```

Using these gene score values, we can assign cluster identities: 

```{r}
crc <- RenameIdents(
  object = crc,
  "0" = "Epithelial",
  "1" = "Myeloid_1",
  "2" = "Epithelial",
  "3" = "Epithelial",
  "4" = "Basophil",
  "5" = "Basophil",
  "6" = "Myeloid_2",
  "7" = "Tcell",
  "8" = "Epithelial"
)
```

One of the myeloid clusters has a lower percentage of fragments in peaks, as
well as a lower overall mitochondrial sequencing depth and a different
nucleosome banding pattern.

```{r cell_filtering_recap, message=FALSE, warning=FALSE}
p1 <- FeatureScatter(crc, "mtDNA_depth", "pct_reads_in_peaks") + ggtitle("") + scale_x_log10()
p2 <- FeatureScatter(crc, "mtDNA_depth", "Nucleosome_signal") + ggtitle("") + scale_x_log10()

p1 + p2 + plot_layout(guides = "collect")
```

We can see that most of the low FRIP cells were the `myeloid 1` cluster. This is
most likely an intra-tumor granulocyte that has relatively poor accessible
chromatin enrichment. Similarly, the unusual nuclear chromatin packaging of this
cell type yields slightly reduced mtDNA coverage compared to the `myeloid 2`
cluster.

## Find mgatk informative mtDNA variants

Next, we can identify sites in the mitochondrial genome that vary across cells, 
and cluster the cells into clonotypes based on the frequency of these variants
in the cells. Signac utilizes the principles established in the original
mtscATAC-seq work of identifying high-quality variants.  

```{r call_variants, message=FALSE, warning=FALSE}
variable.sites <- IdentifyVariants(crc, 
                                   assay = "mgatk", 
                                   refallele = mgatk.data$refallele)
VariantPlot(variants = variable.sites)
```

The plot above clearly shows a group of variants with a higher VMR and strand
concordance. In principle, a high strand concordance reduces the likelihood of
the allele frequency being driven by sequencing error (which predominately
occurs on one but not the other strand. This is due to the preceding nucleotide 
content and a common error in mtDNA genotyping). On the other hand, variants
that have a high VMR are more likely to be clonal variants as the alternate
alleles tend to aggregate in certain cells rather than be equivalently
dispersed about all cells, which would be indicative of some other artifact. 

We note that variants that have a very low VMR and and very high strand
concordance are homoplasmic variants for this sample. While these may be
interesting in some settings (e.g. donor demultiplexing), for inferring
subclones, these are not particularly useful. 

Based on these thresholds, we can filter out a set of informative
mitochondrial variants that differ across the cells.

```{r look_at_variants, message=FALSE, warning=FALSE}
# Establish a filtered data frame of variants based on this processing
high.conf <- subset(
  variable.sites,
  subset = n_cells_conf_detected >= 5 &
    strand_correlation >= 0.65 &
    vmr > 0.01
)

high.conf[, c(1, 2, 5)]
```

A few things stand out. First, 6 out of the 9 variants occur at less than 1% 
allele frequency in the population. However, 16147C>T is present at about 57.9%.
We'll see that this is a clonal variant marking the epithelial cells.
Additionally, all of the called variants are transitions (A - G or C - T) rather
than transversion mutations (A - T or C - G). This fits what we know about how
these mutations arise in the mitochondrial genome. 

Depending on your analytical question, these thresholds can be adjusted to
identify variants that are more prevalent in other cells. 

## Compute the allele frequency of informative mgatk variants for each cell

We currently have information for each strand stored in the `mgatk` assay to allow
strand concordance to be assessed. Now that we have our set of high-confidence
informative variants, we can create a new assay containing strand-collapsed
allele frequency counts for each cell for these variants using the `AlleleFreq()`
function. 

```{r}
crc <- AlleleFreq(
  object = crc,
  variants = high.conf$variant,
  assay = "mgatk",
  new.assay.name = "mgatk_af"
)
crc[["mgatk_af"]]
```

## Visualize informative mgatk variants

Now that the allele frequencies are stored as an additional assay, we can use
the standard functions in Seurat to visualize how these allele frequencies are
distributed across the cells. Here we visualize a subset of the variants using 
`FeaturePlot()` and `DoHeatmap()`.

```{r visualize variants, fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
DefaultAssay(crc) <- "mgatk_af"
alleles.view <- c("12889G>A", "16147C>T", 
                  "9728C>T", "9804G>A")
FeaturePlot(
  object = crc,
  features = alleles.view,
  order = TRUE,
  cols = c("grey", "darkred"),
  ncol = 4
) & NoLegend()
```

```{r fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
DoHeatmap(crc, features = sort(rownames(crc)), slot = "data", disp.max = 1) +
  scale_fill_viridis_c()
```

Here, we can see a few interesting patterns for the selected variants. 16147C>T
is present in essentially all epithelial cells and almost exclusively in
epithelial cells (the edge cases where this isn't true are also cases where the
UMAP and clustering don't full agree). It is at 100% allele frequency-- strongly
suggestive of whatever cell of origin of this tumor had the mutation at 100% and
then expanded. We then see at least 3 variants 1227G>A, 12889G>A, and 9728C>T that
are mostly present specifically in the epithelial cells that define subclones. 
Other variants including 3244G>A, 9804G>A, and 824T>C are found specifically
in immune cell populations, suggesting that these arose from a common hematopoetic
progenitor cell (probably in the bone marrow).


## Loading MQuad mitochondrial variant data

We next load the AD (alternate allele depth) & DP (sum of alternate + reference 
allele depth) sparse matrices for informative variants identified by
[MQuad](https://github.com/single-cell-genetics/MQuad) using the `ReadMQuad` function. 
Bi-alleloc genotyping of the chrM DNA was performed using
[cellsnp-lite](https://github.com/single-cell-genetics/cellsnp-lite), 
using the same reference as was used in mgatk
([rCRS](https://github.com/caleblareau/mgatk/blob/master/mgatk/bin/anno/fasta/rCRS.fasta)).
MQuad takes the genotyping data as input and utilizes a binomial mixture model 
to assess the heteroplasmy of mtDNA variants from background noise.
We will add the AD counts to the `mquad` assay, and add the DP counts as another
layer in the assay. These matrices will then be used to calculate the allele 
frequency of the informative variants per cell.

```{r import_mquad, message=FALSE, warning=FALSE}
# load MQuad data
mquad.data <- ReadMQuad(dir = file.path(crc_dir, 
                                        "mquad/crc_rCRS"),
                        cb = file.path(crc_dir,
                                       "cellsnp/crc_rCRS/cellSNP.samples.tsv"))

# create an assay with AD counts of filtered cells
passed_cells <- as.vector(colnames(crc))
mquad <- CreateAssay5Object(counts = mquad.data$AD_matrix[,passed_cells])

# Add assay & add DP as layer
crc[["mquad"]] <- mquad
crc[["mquad"]]$DP <- mquad.data$DP_matrix[,passed_cells]
```

## Compute the allele frequency of informative MQuad variants for each cell

Compute MQuad allele frequency & add to data slot of the `mquad` assay. 
Note that there will be some 0 values in the DP matrix, as some variants may not
be detected in some cells. This will result in some NA values in the allele 
frequency. We set these values to 0 in the data slot to facilitate plotting.
However, the biological significance between 0 (all ref) & NA (not detected) 
will get erased, thus it should be interpreted with caution.
```{r mquad_vaf, message=FALSE, warning=FALSE}
# compute VAF
mquad_vaf <- crc[["mquad"]]$counts / crc[["mquad"]]$DP
crc[["mquad"]]$data <- mquad_vaf

# Check number of 0 counts in DP matrix
sum(mquad.data$DP_matrix[,passed_cells] == 0)
sum(is.na(mquad_vaf))
crc[["mquad"]]$data[is.na(crc[["mquad"]]$data)] <- 0
```

## Visualize informative MQuad variants

We can now visualize the subset of the variants identified by both mgatk & MQuad 
using `FeaturePlot()`.
```{r visualize_variants, fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
DefaultAssay(crc) <- "mquad"
alleles.view <- c("12889G>A", "16147C>T", 
                  "9728C>T", "9804G>A")
FeaturePlot(
  object = crc,
  features = alleles.view,
  order = TRUE,
  cols = c("grey", "darkred"),
  ncol = 4
) & NoLegend()
```


```{r fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
DoHeatmap(crc, features = sort(rownames(crc)), slot = "data", disp.max = 1) +
  scale_fill_viridis_c()
```

## Investigate overlap between mgatk & MQuad variants

From the heatmaps, the detected mitochondrial variants between mgatk and MQuad 
mostly overlaps. We will investigate the overlapping and distinct variants further.
```{r message=FALSE, warning=FALSE}
# list of variants by each tool
mgatk_variants <- high.conf$variant
mquad_variants <- rownames(mquad_vaf)

# plot venn
p1 <- ggVennDiagram(x = list(mgatk = mgatk_variants, mquad = mquad_variants),
                   set_size = 3, label_size = 3)
p1 + NoLegend() + coord_fixed(clip = "off")
```

A total of 7 variants overlap between mgatk and MQuad, with 2 distinct variants
identified by each tool separately. 

## Distinct variants identified by MQuad

We now investigate the distinct variants by plotting them with `FeaturePlot()`.
```{r fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
# get distinct MQuad variants
mquad_unique <- setdiff(mquad_variants, mgatk_variants)

DefaultAssay(crc) <- "mquad"
FeaturePlot(
  object = crc,
  features = mquad_unique,
  order = TRUE,
  cols = c("grey", "darkred"),
  ncol = 2
) & NoLegend()
```

We can also visualize the 2 distinct variants identified by MQuad on the 
`VariantPlot()` to see where they are in the VMR-strand concordance space.
```{r message=FALSE, warning=FALSE}
p2 <- VariantPlot(variable.sites)

p2$data$highlight <- p2$data$variant %in% mquad_unique

p2 + 
  geom_point(aes(color = highlight)) +
  scale_color_manual(values = c("grey", "blue")) +
  geom_text_repel(
    data = subset(p$data, variant %in% mquad_unique),
    aes(label = variant),
    color = "blue"
  )
```


From the plot, the variant 310T>C was filtered out from mgatk as it did not 
pass the strand concordance test. Variant 16093T>C passed both VMR & strand 
concordance, but was not confidently detected in >=5 cells.
```{r message=FALSE, warning=FALSE}
mquad_variable.sites <- variable.sites[variable.sites$variant %in% mquad_unique,]

mquad_variable.sites[, c(4:7)]
```


## Distinct variants identified by mgatk

We next investigate mgatk distinct variants by plotting them with `FeaturePlot()`.
```{r fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
# get distinct mgatk variants
mgatk_unique <- setdiff(mgatk_variants, mquad_variants)

DefaultAssay(crc) <- "mgatk_af"
FeaturePlot(
  object = crc,
  features = mgatk_unique,
  order = TRUE,
  cols = c("grey", "darkred"),
  ncol = 2
) & NoLegend()
```
The above 2 variants were filtered out by MQuad as they did not pass the cutoff 
in ΔBIC (Bayesian Information Criterion), and thus were not thought to be 
clonally informative. 

Overall, both mitochondrial variant tools show high concordance, and either 
tool can be used for clonal analysis.


# TF1 cell line dataset

Next we'll demonstrate a similar workflow to identify cell clones in a different
dataset, this time generated from a TF1 cell line. This dataset contains more
clones present at a higher proportion, based on the experimental design.

We'll demonstrate how to identify groups of related cells (clones) by clustering
the allele frequency data and how to relate these clonal groups to 
accessibility differences utilizing the multimodal capabilities of Signac. 

## Data loading

<details>
  <summary>**View data download code**</summary>

To download the data from Zenodo run the following in a shell:

```{bash eval=FALSE}
# ATAC data
wget https://zenodo.org/record/3977808/files/TF1.filtered.fragments.tsv.gz
wget https://zenodo.org/record/3977808/files/TF1.filtered.fragments.tsv.gz.tbi
wget https://zenodo.org/record/3977808/files/TF1.filtered.narrowPeak.gz

# mitochondrial genome data
wget https://zenodo.org/record/3977808/files/TF1_filtered.A.txt.gz
wget https://zenodo.org/record/3977808/files/TF1_filtered.T.txt.gz
wget https://zenodo.org/record/3977808/files/TF1_filtered.C.txt.gz
wget https://zenodo.org/record/3977808/files/TF1_filtered.G.txt.gz
wget https://zenodo.org/record/3977808/files/TF1_filtered.chrM_refAllele.txt.gz
wget https://zenodo.org/record/3977808/files/TF1_filtered.depthTable.txt.gz
```

</details>

```{r importData, message=FALSE, warning=FALSE}
# load counts and metadata from cellranger-atac
tf1_dir <- ("/charonfs/scratch/users/astar/gis/leungjy1/Projects/tf1_mitoatac")
counts.path <- file.path(tf1_dir, "10XCR/tf1_1/outs/filtered_peak_bc_matrix.h5")
counts <- Read10X_h5(filename = counts.path)

metadata <- read.csv(
  file = file.path(tf1_dir, "10XCR/tf1_1/outs/singlecell.csv"),
  header = TRUE,
  row.names = 1
)

# create object
tf1_assay <- CreateGRangesAssay(
  counts = counts,
  annotation = annotations,
  min.cells = 10,
  fragments = file.path(tf1_dir, "10XCR/tf1_1/outs/fragments.tsv.gz")
)

tf1 <- CreateSeuratObject(
  counts = tf1_assay,
  assay = "peaks",
  meta.data = metadata
)

tf1[["peaks"]]

# Subset to tf1 cells
tf1_bc <- read.delim(file.path(tf1_dir, "tf1_1_subset_bc.tsv"),
                     sep = '\t',
                     header = FALSE)

tf1 <- subset(tf1, cells = tf1_bc[,1])
tf1
```


## Quality control

```{r message=FALSE, warning=FALSE}
DefaultAssay(tf1) <- "peaks"

# compute TSS enrichment score and nucleosome banding pattern
tf1 <- ATACqc(object = tf1, fragtk.path = Sys.which('fragtk'))
```

```{r}
VlnPlot(tf1, c("nCount_peaks", "Nucleosome_signal", "TSS_enrichment"), pt.size = 0.1)
```

```{r}
tf1 <- subset(
  x = tf1,
  subset = nCount_peaks > 500 &
    Nucleosome_signal < 1 &
    TSS_enrichment > 3
)
tf1
```

```{r}
# read mgatk data
mgatk.data <- ReadMGATK(dir = file.path(tf1_dir, "mgatk_1/final"))

mgatk <- CreateAssayObject(counts = mgatk.data$counts)

# Subset to cell present in the scATAC-seq assat
mgatk <- subset(mgatk, cells = Cells(tf1))

# add assay and metadata to the seurat object
tf1[["mgatk"]] <- mgatk
tf1 <- AddMetaData(tf1, metadata = mgatk.data$depth[Cells(mgatk), ], 
                   col.name = "mtDNA_depth")
```


```{r fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
VlnPlot(tf1, "mtDNA_depth", pt.size = 0.1) + scale_y_log10()
```

```{r}
tf1 <- subset(tf1, mtDNA_depth >=10)
tf1
```


## Identifying mgatk variants

```{r}
DefaultAssay(tf1) <- "mgatk"
variants <- IdentifyVariants(tf1, refallele = mgatk.data$refallele)
VariantPlot(variants)
```

```{r}
high.conf <- subset(
  variants,
  subset = n_cells_conf_detected >= 5 &
    strand_correlation >= 0.65 &
    vmr > 0.01
)
```

```{r}
tf1 <- AlleleFreq(tf1, 
                  variants = high.conf$variant, 
                  assay = "mgatk",
                  new.assay.name = "mgatk_af")
tf1[["mgatk_af"]]
```

## Identifying mgatk clones

Now that we've identified a set of variable alleles, we can cluster the cells
based on the frequency of each of these alleles using the `FindClonotypes()`
function.

```{r message=FALSE, warning=FALSE}
DefaultAssay(tf1) <- "mgatk_af"
tf1 <- FindClonotypes(tf1)
```

```{r}
table(Idents(tf1))
```

Here we see that the clonal clustering has identified 10 different clones in the
TF1 dataset. We can further visualize the frequency of alleles in these clones
using `DoHeatmap()`. The `FindClonotypes()` function also performs hierarchical
clustering on both the clonotypes and the alleles, and sets the factor levels
for the clonotypes based on the hierarchical clustering order, and the order of
variable features based on the hierarchical feature clustering. This allows us
to get a decent ordering of both features and clones automatically:

```{r fig.width=8, fig.height=7, message=FALSE, warning=FALSE}
DoHeatmap(tf1, features = VariableFeatures(tf1), slot = "data", disp.max = 0.3) +
  scale_fill_viridis_c()
```

## Find differentially accessible peaks between mgatk clones

Next we can use the clonal information derived from the mitochondrial assay
to find peaks that are differentially accessible between clones.

```{r message=FALSE, warning=FALSE}
DefaultAssay(tf1) <- "peaks"

# find peaks specific to one clone
markers.fast <- FoldChange(tf1, ident.1 = 2, slot = "counts")
markers.fast <- markers.fast[order(markers.fast$avg_log2FC, decreasing = TRUE), ] # sort by fold change
head(markers.fast)
```

We can visualise the DNA accessibility in these regions for each clone using
the `CoveragePlot()` function. As you can see, the peaks identified are highly
specific to one clone.

```{r}
CoveragePlot(
  object = tf1,
  region = rownames(markers.fast)[1],
  extend.upstream = 2000,
  extend.downstream = 2000
)
```


## MQuad variants & clonotypes

```{r message=FALSE, warning=FALSE}
mquad.data <- ReadMQuad(dir = file.path(tf1_dir, "mquad/tf1_1_rCRS"), 
                        cb = file.path(tf1_dir, "cellsnp/tf1_1_rCRS/cellSNP.samples.tsv"))

# create an assay with AD counts of filtered cells
passed_cells <- as.vector(colnames(tf1))
mquad <- CreateAssay5Object(counts = mquad.data$AD_matrix[,passed_cells])

# Add assay & add DP as layer
tf1[["mquad"]] <- mquad
tf1[["mquad"]]$DP <- mquad.data$DP_matrix[,passed_cells]
```

Compute MQuad VAF & add to data slot of mquad assay. 
```{r mquad_vaf, message=FALSE, warning=FALSE}
# compute VAF
mquad_vaf <- tf1[["mquad"]]$counts / tf1[["mquad"]]$DP
tf1[["mquad"]]$data <- mquad_vaf

# Check number of 0 counts in DP matrix
sum(mquad.data$DP_matrix[,passed_cells] == 0)
sum(is.na(mquad_vaf))
# Set NAs in VAF to 0 
tf1[["mquad"]]$data[is.na(tf1[["mquad"]]$data)] <- 0
```


We perform clonotype clustering using the identified MQuad variants.
```{r message=FALSE, warning=FALSE}
DefaultAssay(tf1) <- "mquad"
tf1 <- FindClonotypes(tf1, assay = "mquad")
table(Idents(tf1))
```

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
DoHeatmap(tf1, features = VariableFeatures(tf1), slot = "data", disp.max = 0.3) +
  scale_fill_viridis_c()
```


## Investigate overlap between mgatk & MQuad variants

```{r message=FALSE, warning=FALSE}
# list of variants by each tool
mgatk_variants <- high.conf$variant
mquad_variants <- rownames(mquad_vaf)

# plot venn
p3 <- ggVennDiagram(x = list(mgatk = mgatk_variants, mquad = mquad_variants),
                   set_size = 3, label_size = 3)
p3 + NoLegend() + coord_fixed(clip = "off")
```


<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
