---
title: "Motif analysis with Signac"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this tutorial, we will perform DNA sequence motif analysis in Signac. We will
explore two complementary options for performing motif analysis: one by finding
overrepresented motifs in a set of differentially accessible peaks, one method
performing differential motif activity analysis between groups of cells.

In this demonstration we use data from the adult mouse brain. See our
[vignette](mouse_brain_vignette.html) for the code used to generate this object,
and links to the raw data. First, load the required packages and the
pre-computed Seurat object:

```{r include=FALSE}
if (!requireNamespace("TFBSTools", quietly = TRUE))
    BiocManager::install("TFBSTools")

if (!requireNamespace("JASPAR2020", quietly = TRUE))
    BiocManager::install("JASPAR2020")

if (!requireNamespace("BSgenome.Mmusculus.UCSC.mm10", quietly = TRUE))
    BiocManager::install("BSgenome.Mmusculus.UCSC.mm10")

if (!requireNamespace("chromVAR", quietly = TRUE))
    BiocManager::install("chromVAR")

if (!requireNamespace("ggbio", quietly = TRUE))
    BiocManager::install("ggbio")
```

```{r message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)
library(JASPAR2020)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10)
library(patchwork)
set.seed(1234)
```

```{r message=FALSE, warning=FALSE}
mouse_brain <- readRDS("../vignette_data/adult_mouse_brain.rds")
mouse_brain
```

```{r message=FALSE, warning=FALSE}
p1 <- DimPlot(mouse_brain, label = TRUE, pt.size = 0.1) + NoLegend()
p1
```

## Adding motif information to the Seurat object

To add the DNA sequence motif information required for motif analyses, we can
run the `AddMotifs()` function:

```{r message=FALSE, warning=FALSE}
# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)

# add motif information
mouse_brain <- AddMotifs(
  object = mouse_brain,
  genome = BSgenome.Mmusculus.UCSC.mm10,
  pfm = pfm
)
```

To facilitate motif analysis in Signac, we have create the `Motif` class to
store all the required information, including a list of position weight matrices
(PWMs) or position frequency matrices (PFMs) and a motif occurrence matrix.
Here, the `AddMotifs()` function construct a `Motif` object and adds it to our
mouse brain dataset, along with other information such as the base composition
of each peak. A motif object can be added to any Seurat assay using the `SetAssayData()`
function. See the [object interaction vignette](data_structures.html) for
more information.

## Finding overrepresented motifs

To identify potentially important cell-type-specific regulatory sequences, we
can search for DNA motifs that are overrepresented in a set of peaks that are
differentially accessible between cell types.

Here, we find differentially accessible peaks between Pvalb and Sst inhibitory
interneurons. For sparse data (such as scATAC-seq), we find it is often necessary
to lower the `min.pct` threshold in `FindMarkers()` from the default (0.1, which
was designed for scRNA-seq data). 

We then perform a hypergeometric test to test the probability of
observing the motif at the given frequency by chance, comparing with a
background set of peaks matched for GC content. 

```{r message=FALSE, warning=FALSE}
da_peaks <- FindMarkers(
  object = mouse_brain,
  ident.1 = 'Pvalb',
  ident.2 = 'Sst',
  only.pos = TRUE,
  test.use = 'LR',
  min.pct = 0.05,
  latent.vars = 'nCount_peaks'
)

# get top differentially accessible peaks
top.da.peak <- rownames(da_peaks[da_peaks$p_val < 0.005, ])
```

<details>
  <summary>**Optional: choosing a set of background peaks**</summary>
  
Matching the set of background peaks is essential when finding enriched DNA
sequence motifs. By default, we choose a set of peaks matched for GC content,
but it can be sometimes be beneficial to further restrict the background peaks
to those that are accessible in the groups of cells compared when finding
differentially accessible peaks.

The `AccessiblePeaks()` function can be used to find a set of peaks that are 
open in a subset of cells. We can use this function to first restrict the set
of possible background peaks to those peaks that were open in the set of cells
compared in `FindMarkers()`, and then create a GC-content-matched set of peaks
from this larger set using `MatchRegionStats()`.

```{r}
# find peaks open in Pvalb or Sst cells
open.peaks <- AccessiblePeaks(mouse_brain, idents = c("Pvalb", "Sst"))

# match the overall GC content in the peak set
meta.feature <- GetAssayData(mouse_brain, assay = "peaks", slot = "meta.features")
peaks.matched <- MatchRegionStats(
  meta.feature = meta.feature[open.peaks, ],
  query.feature = meta.feature[top.da.peak, ],
  n = 50000
)
```

`peaks.matched` can then be used as the background peak set by setting
`background=peaks.matched` in `FindMotifs()`.

</details>

```{r}
# test enrichment
enriched.motifs <- FindMotifs(
  object = mouse_brain,
  features = top.da.peak
)
```

```{r echo=FALSE}
knitr::kable(head(enriched.motifs))
```

We can also plot the position weight matrices for the motifs, so we can
visualize the different motif sequences.

```{r}
MotifPlot(
  object = mouse_brain,
  motifs = head(rownames(enriched.motifs))
)
```

We and others have previously shown that Mef-family motifs, particularly
*Mef2c*, are enriched in Pvalb-specific peaks in scATAC-seq data
(https://doi.org/10.1016/j.cell.2019.05.031; https://doi.org/10.1101/615179),
and further shown that *Mef2c* is required for the development of Pvalb
interneurons (https://www.nature.com/articles/nature25999). Here our results are 
consistent with these findings, and we observe a strong enrichment of Mef-family
motifs in the top results from `FindMotifs()`.

## Computing motif activities

We can also compute a per-cell motif activity score by running 
[chromVAR](https://greenleaflab.github.io/chromVAR/index.html). This allows us
to visualize motif activities per cell, and also provides an alternative method
of identifying differentially-active motifs between cell types.

ChromVAR identifies motifs associated with variability in chromatin
accessibility between cells. See the chromVAR
[paper](https://www.nature.com/articles/nmeth.4401) for a complete description
of the method.

```{r message=FALSE, warning=FALSE, fig.width=12}
mouse_brain <- RunChromVAR(
  object = mouse_brain,
  genome = BSgenome.Mmusculus.UCSC.mm10
)

DefaultAssay(mouse_brain) <- 'chromvar'

# look at the activity of Mef2c
p2 <- FeaturePlot(
  object = mouse_brain,
  features = "MA0497.1",
  min.cutoff = 'q10',
  max.cutoff = 'q90',
  pt.size = 0.1
)
p1 + p2
```

We can also directly test for differential activity scores between cell types.
This tends to give similar results as performing an enrichment test on
differentially accessible peaks between the cell types (shown above).

When performing differential testing on the chromVAR z-score, we can set 
`mean.fxn=rowMeans` and `fc.name="avg_diff"` in the `FindMarkers()` function so
that the fold-change calculation computes the average difference in z-score
between the groups.

```{r message=FALSE, warning=FALSE}
differential.activity <- FindMarkers(
  object = mouse_brain,
  ident.1 = 'Pvalb',
  ident.2 = 'Sst',
  only.pos = TRUE,
  mean.fxn = rowMeans,
  fc.name = "avg_diff"
)

MotifPlot(
  object = mouse_brain,
  motifs = head(rownames(differential.activity)),
  assay = 'peaks'
)
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>

