---
title: "Analyzing adult mouse brain scATAC-seq"
output: html_document
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "../vignette_data/mouse_brain")
```

For this tutorial, we will be analyzing a single-cell ATAC-seq dataset of adult
mouse brain cells provided by 10x Genomics. The following files are used in this
vignette, all available through the 10x Genomics website.

<details>
  <summary>**View data download code**</summary>

To download the required data, run the following lines in a shell:

```{bash eval=FALSE}
wget http://cf.10xgenomics.com/samples/cell-atac/1.1.0/atac_v1_adult_brain_fresh_5k/atac_v1_adult_brain_fresh_5k_filtered_peak_bc_matrix.h5
wget http://cf.10xgenomics.com/samples/cell-atac/1.1.0/atac_v1_adult_brain_fresh_5k/atac_v1_adult_brain_fresh_5k_singlecell.csv
wget http://cf.10xgenomics.com/samples/cell-atac/1.1.0/atac_v1_adult_brain_fresh_5k/atac_v1_adult_brain_fresh_5k_fragments.tsv.gz
wget http://cf.10xgenomics.com/samples/cell-atac/1.1.0/atac_v1_adult_brain_fresh_5k/atac_v1_adult_brain_fresh_5k_fragments.tsv.gz.tbi
```

</details>

This vignette echoes the commands run in the introductory Signac vignette on
[human PBMC](pbmc_vignette.html). We
provide the same analysis in a different system to demonstrate performance and
applicability to other tissue types, and to provide an example from another
species.

First load in Signac, Seurat, and some other packages we will be using for
analyzing mouse data.

```{r include=FALSE}
if (!requireNamespace("EnsDb.Mmusculus.v79", quietly = TRUE)) {
  BiocManager::install("EnsDb.Mmusculus.v79")
}
```

```{r setup, message=FALSE}
library(Signac)
library(Seurat)
library(EnsDb.Mmusculus.v79)
library(ggplot2)
library(patchwork)
```

## Pre-processing workflow

```{r}
counts <- Read10X_h5("atac_v1_adult_brain_fresh_5k_filtered_peak_bc_matrix.h5")
metadata <- read.csv(
  file = "atac_v1_adult_brain_fresh_5k_singlecell.csv",
  header = TRUE,
  row.names = 1
)

brain_assay <- CreateGRangesAssay(
  counts = counts,
  sep = c(":", "-"),
  fragments = "atac_v1_adult_brain_fresh_5k_fragments.tsv.gz",
  min.cells = 1
)
brain <- CreateSeuratObject(
  counts = brain_assay,
  assay = "peaks",
  project = "ATAC",
  meta.data = metadata
)
```

We can also add gene annotations to the `brain` object for the mouse genome.
This will allow downstream functions to pull the gene annotation information
directly from the object.

```{r message=FALSE, warning=FALSE}
# extract gene annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)

# change to UCSC style since the data was mapped to mm10
seqlevels(annotations) <- paste0("chr", seqlevels(annotations))
genome(annotations) <- "mm10"

# add the gene information to the object
Annotation(brain) <- annotations
```

## Computing QC Metrics

Next we compute some useful per-cell QC metrics including the nucleosome signal,
TSS enrichment score, and total counts per cell using the `ATACqc()` function:

```{r message=FALSE, warning=FALSE}
brain <- ATACqc(brain)
```

```{r message=FALSE, warning=FALSE, fig.width=18, fig.height=6}
VlnPlot(
  object = brain,
  features = c("total_fragments", "TSS_enrichment", "Nucleosome_signal", "FRiP"),
  pt.size = 0.1,
  ncol = 5
)
```

We remove cells that are outliers for these QC metrics.

```{r}
brain <- subset(
  x = brain,
  subset = total_fragments > 3000 &
    total_fragments < 100000 &
    Nucleosome_signal < 4 &
    TSS_enrichment > 7
)
brain
```

## Normalization and linear dimensional reduction

```{r message=FALSE, warning=FALSE}
brain <- RunTFIDF(brain)
brain <- FindTopFeatures(brain, min.cutoff = "q0")
brain <- RunSVD(object = brain)
```

The first LSI component often captures sequencing depth (technical variation)
rather than biological variation. If this is the case, the component should be
removed from downstream analysis. We can assess the correlation between each LSI
component and sequencing depth using the `DepthCor()` function:

```{r}
DepthCor(brain)
```

Here we see there is a very strong correlation between the first LSI component
and the total number of counts for the cell, so we will perform downstream steps
without this component.

## Non-linear dimension reduction and clustering

Now that the cells are embedded in a low-dimensional space, we can use methods
commonly applied for the analysis of scRNA-seq data to perform graph-based
clustering, and non-linear dimension reduction for visualization. The functions
`RunUMAP()`, `FindNeighbors()`, and `FindClusters()` all come from the Seurat
package.

```{r message=FALSE, warning=FALSE}
brain <- RunUMAP(
  object = brain,
  reduction = "lsi",
  dims = 2:30
)
brain <- FindNeighbors(
  object = brain,
  reduction = "lsi",
  dims = 2:30
)
brain <- FindClusters(
  object = brain,
  algorithm = 3,
  resolution = 1.2,
  verbose = FALSE
)

DimPlot(object = brain, label = TRUE) + NoLegend()
```

## Create a gene activity matrix

```{r message=FALSE, warning=FALSE}
# compute gene activities
gene.activities <- GeneActivity(brain)

# add the gene activity matrix to the Seurat object as a new assay
brain[["RNA"]] <- CreateAssayObject(counts = gene.activities)
brain <- NormalizeData(brain, assay = "RNA")
```

```{r fig.width=12, fig.height=10}
DefaultAssay(brain) <- "RNA"
FeaturePlot(
  object = brain,
  features = c("Sst", "Pvalb", "Gad2", "Neurod6", "Rorb", "Syt6"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 3
)
```

## Integrating with scRNA-seq data

To help interpret the scATAC-seq data, we can classify cells based on an
scRNA-seq experiment from the same biological system (the adult mouse brain).
We utilize methods for cross-modality integration and label transfer, described
[here](https://doi.org/10.1016/j.cell.2019.05.031), with a more in-depth
tutorial [here](https://satijalab.org/seurat/v3.0/atacseq_integration_vignette.html).

You can download the raw data for this experiment from the Allen Institute
[website](http://celltypes.brain-map.org/api/v2/well_known_file_download/694413985),
and view the code used to construct this object on
[GitHub](https://github.com/satijalab/Integration2019/blob/master/preprocessing_scripts/allen_brain.R). 
Alternatively, you can download the pre-processed Seurat object
[here](https://signac-objects.s3.amazonaws.com/allen_brain.rds).

```{r warning=FALSE, message=FALSE}
# Load the pre-processed scRNA-seq data
allen_rna <- readRDS("allen_brain.rds")
allen_rna <- UpdateSeuratObject(allen_rna)
allen_rna <- FindVariableFeatures(
  object = allen_rna,
  nfeatures = 5000
)

transfer.anchors <- FindTransferAnchors(
  reference = allen_rna,
  query = brain,
  reduction = "cca",
  dims = 1:30
)

predicted.labels <- TransferData(
  anchorset = transfer.anchors,
  refdata = allen_rna$subclass,
  weight.reduction = brain[["lsi"]],
  dims = 2:30
)

brain <- AddMetaData(object = brain, metadata = predicted.labels)
```


```{r fig.width=12}
plot1 <- DimPlot(allen_rna, group.by = "subclass", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("scRNA-seq")
plot2 <- DimPlot(brain, group.by = "predicted.id", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("scATAC-seq")
plot1 + plot2
```

You can see that the RNA-based classifications are consistent with the
UMAP visualization, computed only on the ATAC-seq data.

## Find differentially accessible peaks between clusters

Here, we find differentially accessible regions between excitatory neurons in
different layers of the cortex.

```{r message=TRUE, warning=FALSE}
# switch back to working with peaks instead of gene activities
DefaultAssay(brain) <- "peaks"
Idents(brain) <- "predicted.id"

da_peaks <- FindMarkers(
  object = brain,
  ident.1 = c("L2/3 IT"),
  ident.2 = c("L4", "L5 IT", "L6 IT"),
  test.use = "wilcox",
  min.pct = 0.1
)

head(da_peaks)
```

```{r fig.width=12}
plot1 <- VlnPlot(
  object = brain,
  features = rownames(da_peaks)[1],
  pt.size = 0.1,
  idents = c("L4", "L5 IT", "L2/3 IT")
)
plot2 <- FeaturePlot(
  object = brain,
  features = rownames(da_peaks)[1],
  pt.size = 0.1,
  max.cutoff = "q95"
)
plot1 | plot2
```

## Plotting genomic regions

We can also create coverage plots grouped by cluster, cell type, or any other
metadata stored in the object for any genomic region using the `CoveragePlot()`
function. These represent pseudo-bulk accessibility tracks, where signal from
all cells within a group have been averaged together to visualize the DNA 
accessibility in a region.

```{r message=FALSE, warning=FALSE, out.width="90%", fig.height=10}
# show cell types with at least 50 cells
idents.plot <- names(which(table(Idents(brain)) > 50))
brain <- SortIdents(brain)

CoveragePlot(
  object = brain,
  region = c("Neurod6", "Gad2"),
  idents = idents.plot,
  extend.upstream = 1000,
  extend.downstream = 1000,
  ncol = 1
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qs2::qs_save(object = brain, file = "adult_mouse_brain.qs2", nthreads = 4)
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
