---
title: "Analyzing PBMC scATAC-seq"
output: html_document
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "../vignette_data/pbmc_vignette")
```

```{r, include=FALSE}
options(width = 600)
local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) {
      options$attr.output <- c(
        options$attr.output,
        sprintf('style="max-height: %s;"', options$max.height)
      )
    }
    hook_output(x, options)
  })
})
```

For this tutorial, we will be analyzing a single-cell ATAC-seq dataset of human
peripheral blood mononuclear cells (PBMCs) provided by 10x Genomics. We will use
the set of regulatory element modules (REMO) that we have recently released for 
the human genome. By using REMO rather than *de novo* peak calling, we can skip 
the peak calling step and ensure that our dataset can be easily compared to
other datasets. We can also use REMO to perform some automated cell annotation,
demonstrated at the end section of this vignette.

REMO can be installed from GitHub:

```{r eval=FALSE, include=TRUE}
remotes::install_github("stuart-lab/REMO.v1.GRCh38")
```

<details>
  <summary>**View data download code**</summary>

To download all the required files, you can run the following lines in a shell:

```{sh eval=FALSE}
wget https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/10k_pbmc_ATACv2_nextgem_Chromium_Controller_filtered_peak_bc_matrix.h5
wget https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/10k_pbmc_ATACv2_nextgem_Chromium_Controller_singlecell.csv
wget https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz
wget https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz.tbi
```

</details>

First load in Signac, Seurat, and some other packages we will be using for
analyzing human data.

```{r setup, message=FALSE}
library(Signac)
library(Seurat)
library(REMO.v1.GRCh38)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
```

## Pre-processing workflow

When pre-processing chromatin data, Signac uses the ATAC fragment file as the
main input data. This represents a full list of all unique fragments across all
single cells. It is stored on-disk for memory efficiency and scalability. More
information about the fragment file can be found on the 10x Genomics [website](https://support.10xgenomics.com/single-cell-atac/software/pipelines/latest/output/fragments) or on the [sinto website](https://timoast.github.io/sinto/basic_usage.html#create-scatac-seq-fragments-file).

We start by selecting cells to include in the analysis:

```{r warning=FALSE, message=FALSE}
# load gene annotations
library(AnnotationHub)
ah <- AnnotationHub()

ensdb_v98 <- ah[["AH75011"]]
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

# change to UCSC style since the data was mapped to hg38
seqlevelsStyle(annotations) <- "UCSC"

frags <- "10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz"
metadata <- ATACqc(object = frags, annotations = annotations)
head(metadata)
```

```{r}
cells <- rownames(metadata[metadata$TSS_enrichment > 5 & metadata$total_fragments > 8000, ])
```

Create REMO matrix using [fragtk](https://crates.io/crates/fragtk):

```{r include=FALSE}
if (!requireNamespace("REMO.v1.GRCh38", quietly = TRUE)) {
  remotes::install_github(repo = "stuart-lab/REMO.v1.GRCh38")
}
```

```{r}
remo <- RunFragtk(
  fragments = frags,
  features = REMO.v1.GRCh38,
  cells = cells,
  group = TRUE
)
```

Create Seurat object

```{r}
chrom_assay <- CreateChromatinAssay5(
  counts = remo,
  fragments = "10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz",
  annotation = annotations,
  min.cells = 20
)

pbmc <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "REMO",
  meta.data = metadata,
  min.cells = 20
)
```

```{r}
pbmc
```

We can also add gene annotations to the `pbmc` object for the human genome.
This will allow downstream functions to pull the gene annotation information
directly from the object.

Multiple patches are released for each genome assembly. When dealing with mapped 
data (such as the 10x Genomics files we will be using), it is advisable to use 
the annotations from the same assembly patch that was used to perform the mapping.

From the [dataset summary](https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/10k_pbmc_ATACv2_nextgem_Chromium_Controller_web_summary.html), 
we can see that the reference package 10x Genomics used to perform 
the mapping was “GRCh38-2020-A”, which [corresponds to](https://www.10xgenomics.com/support/software/cell-ranger/downloads/cr-ref-build-steps#ref-2020-a) the Ensembl v98 patch release. 
For more information on the various Ensembl releases, you can refer to [this site](https://www.ensembl.org/info/website/archives/assembly.html).

## Quality control

We can compute quality control metrics for the scATAC-seq data using the `ATACqc` function.
These metrics include:

* Nucleosome banding pattern: The histogram of DNA fragment sizes (determined from the paired-end sequencing reads) should exhibit a strong nucleosome banding pattern corresponding to the length of DNA wrapped around a single nucleosome. We calculate this per single cell, and quantify the approximate ratio of mononucleosomal to nucleosome-free fragments (stored as `Nucleosome_signal`)

* Transcriptional start site (TSS) enrichment score: The [ENCODE project](https://www.encodeproject.org/) has defined an ATAC-seq targeting score based on the ratio of fragments centered at the TSS to fragments in TSS-flanking regions (see https://www.encodeproject.org/data-standards/terms/). Poor ATAC-seq experiments typically will have a low TSS enrichment score. Results are stored in metadata under the column name `TSS_enrichment`.

* Total number of fragments: A measure of cellular sequencing depth / complexity. Cells with very few fragments may need to be excluded due to low sequencing depth. Cells with extremely high levels may represent doublets, nuclei clumps, or other artefacts. This is stored as
`total_fragments`

```{r}
VlnPlot(
  object = pbmc,
  features = c("TSS_enrichment", "Nucleosome_signal", "total_fragments"),
  group.by = "orig.ident",
  ncol = 3,
  pt.size = 0
)
```

The relationship between variables stored in the object metadata can
be visualized using the `DensityScatter()` function. This can also be used
to quickly find suitable cutoff values for different QC metrics by setting
`quantiles=TRUE`:

```{r}
DensityScatter(
  object = pbmc,
  x = "total_fragments",
  y = "TSS_enrichment",
  log_x = TRUE,
  quantiles = TRUE
)
```


<details>
  <summary>**Doublet removal (optional)**</summary>
  
We can predict doublets using the `scDblFinder` package:

```{r include=FALSE}
if (!requireNamespace("scDblFinder", quietly = TRUE)) {
  BiocManager::install("scDblFinder")
}
```

```{r message=FALSE, warning=FALSE}
library(scDblFinder)

# doublet score
scDbl <- as.SingleCellExperiment(pbmc)
scDbl <- scDblFinder(scDbl)

pbmc$dbl.class <- scDbl$scDblFinder.class
pbmc$dbl.score <- scDbl$scDblFinder.score
```

</details>

```{r}
pbmc <- subset(pbmc, total_fragments > 10000 &
  total_fragments < 60000 &
  TSS_enrichment > 9 &
  dbl.class == "singlet")
```

```{r}
pbmc
```

## Feature selection

We can select highly variable REMO modules by modelling the mean-variance relationship.
Here we use the `FitMeanVar()` function in Signac, which will fit a loess curve to the
log(mean) and log(variance) of each feature, and then estimate the residual variance for
each feature given the mean. This is similar to how the `FindVariableFeatures()` function
in Seurat works, but is more efficient for a large number of features as we downsample
the features when fitting the loess model.

```{r}
pbmc <- FitMeanVar(pbmc)
```

## Normalization and linear dimensional reduction

```{r}
pbmc <- NormalizeData(pbmc)
pbmc <- RunSVD(pbmc, pca = TRUE)
```

## Non-linear dimension reduction and clustering

Now that the cells are embedded in a low-dimensional space we can use methods
commonly applied for the analysis of scRNA-seq data to perform graph-based
clustering and non-linear dimension reduction for visualization. The functions
`RunUMAP()`, `FindNeighbors()`, and `FindClusters()` all come from the Seurat
package.

```{r message=FALSE, warning=FALSE}
pbmc <- RunUMAP(object = pbmc, dims = 1:40)
pbmc <- FindNeighbors(object = pbmc, dims = 1:40)
pbmc <- FindClusters(object = pbmc, verbose = FALSE, algorithm = 3)
DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

## Create a gene activity matrix

The UMAP visualization reveals the presence of multiple cell groups in human 
blood. If you are familiar with 
[scRNA-seq analyses of PBMC](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html),
you may recognize the presence of certain myeloid and lymphoid populations
in the scATAC-seq data. However, annotating and interpreting clusters is more
challenging in scATAC-seq data as much less is known about the functional roles
of noncoding genomic regions than is known about protein coding regions (genes).

We can try to quantify the activity of each gene in the genome by
assessing the chromatin accessibility associated with the gene, and create a
new gene activity assay derived from the scATAC-seq data. Here we will use a
simple approach of summing the fragments intersecting the gene body and promoter
region (we also recommend exploring the
[Cicero](https://cole-trapnell-lab.github.io/cicero-release/) tool, which can
accomplish a similar goal, and we provide a vignette showing how to run Cicero
within a Signac workflow [here](cicero.html)). 

To create a gene activity matrix, we extract gene coordinates and extend them
to include the 2 kb upstream region (as promoter accessibility is often
correlated with gene expression). We then count the number of fragments for each
cell that map to each of these regions, using the using the `FeatureMatrix()`
function. These steps are automatically performed by the `GeneActivity()`
function:

```{r message=FALSE, warning=FALSE}
gene.activities <- GeneActivity(pbmc)
```

```{r message=FALSE, warning=FALSE}
# add the gene activity matrix to the Seurat object as a new assay and normalize it
pbmc[["RNA"]] <- CreateAssay5Object(counts = gene.activities)
pbmc <- NormalizeData(pbmc, assay = "RNA")
```

Now we can visualize the activities of canonical marker genes to help interpret
our ATAC-seq clusters. Note that the activities will be much noisier than
scRNA-seq measurements. This is because they represent measurements from sparse
chromatin data, and because they assume a general correspondence between gene
body/promoter accessibility and gene expression which may not always be the
case. Nonetheless, we can begin to discern populations of monocytes, B, T, and
NK cells based on these gene activity profiles. However, further subdivision of
these cell types is challenging based on supervised analysis alone.

```{r fig.width=12, fig.height=10}
DefaultAssay(pbmc) <- "RNA"

FeaturePlot(
  object = pbmc,
  features = c("MS4A1", "CD3D", "LEF1", "NKG7", "TREM1", "LYZ"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 3
)
```

## Integrating with scRNA-seq data

To help interpret the scATAC-seq data, we can classify cells based on an
scRNA-seq experiment from the same biological system (human PBMC). We utilize
methods for cross-modality integration and label transfer, described 
[here](https://doi.org/10.1016/j.cell.2019.05.031), with a more in-depth
tutorial [here](https://satijalab.org/seurat/v3.0/atacseq_integration_vignette.html).
We aim to identify shared correlation patterns in the gene activity matrix and
scRNA-seq dataset to identify matched biological states across the two
modalities. This procedure returns a classification score for each cell for each
scRNA-seq-defined cluster label.

Here we load a pre-processed scRNA-seq dataset for human PBMCs, also provided by
10x Genomics. You can download the raw data for this experiment from the 10x 
[website](https://support.10xgenomics.com/single-cell-gene-expression/datasets/3.0.0/pbmc_10k_v3),
and view the code used to construct this object on
[GitHub](https://github.com/satijalab/Integration2019/blob/master/preprocessing_scripts/pbmc_10k_v3.R).
Alternatively, you can download the pre-processed Seurat object 
[here](https://signac-objects.s3.amazonaws.com/pbmc_10k_v3.rds).

```{r warning=FALSE, message=FALSE}
# Load the pre-processed scRNA-seq data for PBMCs
pbmc_rna <- readRDS("pbmc_10k_v3.rds")
pbmc_rna <- UpdateSeuratObject(pbmc_rna)
```

```{r}
transfer.anchors <- FindTransferAnchors(
  reference = pbmc_rna,
  query = pbmc,
  reduction = "cca"
)

predicted.labels <- TransferData(
  anchorset = transfer.anchors,
  refdata = pbmc_rna$celltype,
  weight.reduction = pbmc[["pca"]],
  dims = 1:30
)

pbmc <- AddMetaData(object = pbmc, metadata = predicted.labels)
```

```{r fig.width=12}
plot1 <- DimPlot(
  object = pbmc_rna,
  group.by = "celltype",
  label = TRUE,
  repel = TRUE
) + NoLegend() + ggtitle("scRNA-seq")

plot2 <- DimPlot(
  object = pbmc,
  group.by = "predicted.id",
  label = TRUE,
  repel = TRUE
) + NoLegend() + ggtitle("scATAC-seq")

plot1 + plot2
```

You can see that the scRNA-based classifications are consistent with
the UMAP visualization that was computed using the scATAC-seq data. Notice, however, 
that a small population of cells are predicted to be platelets in the scATAC-seq 
dataset. This is unexpected as platelets are not nucleated and should not be detected 
by scATAC-seq. It is possible that the cells predicted to be platelets could 
instead be the platelet-precursors megakaryocytes, which largely reside in the 
bone marrow but are rarely found in the peripheral blood of healthy patients, 
such as the individual these PBMCs were drawn from. Given the already extreme 
rarity of megakaryocytes within normal bone marrow (< 0.1%), this scenario seems unlikely.  

<details>

<summary>**A closer look at the "platelets"**</summary>

Plotting the prediction score for the cells assigned to each label reveals that 
the “platelet” cells received relatively low scores (< 0.8), indicating a low confidence 
in the assigned cell identity. In most cases, the next most likely cell identity 
predicted for these cells was “CD4 naive”.

```{r fig.width=12}
VlnPlot(pbmc, "prediction.score.max", group.by = "predicted.id")
```

```{r max.height='300px'}
# Identify the metadata columns that start with "prediction.score."
metadata_attributes <- colnames(pbmc[[]])
prediction_score_attributes <- grep("^prediction.score.", metadata_attributes, value = TRUE)
prediction_score_attributes <- setdiff(prediction_score_attributes, "prediction.score.max")

# Extract the prediction score attributes for these cells
predicted_platelets <- which(pbmc$predicted.id == "Platelet")
platelet_scores <- pbmc[[]][predicted_platelets, prediction_score_attributes]

# Order the columns by their average values in descending order
ordered_columns <- names(sort(colMeans(platelet_scores, na.rm = TRUE), decreasing = TRUE))
ordered_platelet_scores_df <- platelet_scores[, ordered_columns]

print(ordered_platelet_scores_df)
```
</details>

As there are only a very small number of cells classified as “platelets” (< 20), 
it is difficult to figure out their precise cellular identity. Larger datasets 
would be required to confidently identify specific peaks for this population of
cells, and further analysis performed to correctly annotate them.

For downstream analyses, we can simply reassign the identities of each cell 
from their cluster index to the per-cell predicted labels. It is also 
possible to consider merging the cluster identities and predicted labels.

```{r}
# change cell identities to the per-cell predicted labels
Idents(pbmc) <- pbmc$predicted.id
```

<details>

<summary>**To combine clustering and label transfer results**</summary>

Alternatively, in instances where we wish to combine our scATAC-seq clustering 
and label transfer results, we can reassign the identities of all cells within 
each UMAP cluster to the most common predicted label for that cluster.

```{r eval=FALSE}
# replace each cluster label with its most likely predicted label
for (i in levels(pbmc)) {
  cells_to_reid <- WhichCells(pbmc, idents = i)
  newid <- names(which.max(table(pbmc$predicted.id[cells_to_reid])))
  Idents(pbmc, cells = cells_to_reid) <- newid
}
```

</details>

## Find differentially accessible REMO modules between cell types

To find differentially accessible modules between clusters of cells, we can
perform a differential accessibility (DA) test. A simple approach is to perform
a Wilcoxon rank sum test, and the [presto](https://github.com/immunogenomics/presto)
package has implemented an extremely fast Wilcoxon test that can be run on a
Seurat object.

<details>

<summary>**Install presto**</summary>

```{r eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("immunogenomics/presto")
```

</details>

Here we will focus on comparing Naive CD4 cells and CD14 monocytes, but any groups of
cells can be compared using these methods. We can also visualize these modules
on a violin plot, feature plot, dot plot, heat map, or any
[visualization tool in Seurat](https://satijalab.org/seurat/articles/visualization_vignette.html).

```{r message=TRUE, warning=FALSE}
DefaultAssay(pbmc) <- "REMO"

# wilcox is the default option for test.use
da_remo <- FindMarkers(
  object = pbmc,
  ident.1 = "CD4 Naive",
  ident.2 = "CD14+ Monocytes",
  test.use = "wilcox",
  min.pct = 0.1
)

head(da_remo)
```


```{r, cache=FALSE, fig.width=12}
plot3 <- VlnPlot(
  object = pbmc,
  features = rownames(da_remo)[1],
  pt.size = 0.1,
  idents = c("CD4 Naive", "CD14+ Monocytes")
)
plot4 <- FeaturePlot(
  object = pbmc,
  features = rownames(da_remo)[1],
  pt.size = 0.1
)

plot3 | plot4
```

## Cell type term enrichment

We can also perform an automated cell type annotation using the REMO cell type
ontology terms, through an ontology term enrichment analysis. This is useful if
we don't have access to an annotated reference dataset for label transfer, or
to provide additional cell type predictions independent of reference mapping.

Here we demonstrate how to perform CL term enrichment analysis for each cell
cluster using REMO and the fgsea package.

```{r}
if (!requireNamespace("fgsea", quietly = TRUE)) {
  BiocManager::install("fgsea")
}
```

We can use the `EnrichedTerms()` function in Signac to run ontology enrichment
on a ranked list of features for each cluster. We need to provide a list of
ontology terms associated with each feature in our assay. For REMO, this is
included in the `REMO.v1.GRCh38.terms` object included in the package. We can
subset these terms to only include cell types in a tissue that we're interested
in using the included list of tissue-celltype associations:

```{r}
celltypes <- REMO.v1.GRCh38.terms[REMO.v1.GRCh38.tissues$Blood]
```

```{r message=FALSE, warning=FALSE}
enrichments <- EnrichedTerms(
  object = pbmc,
  terms = celltypes,
  group.by = "seurat_clusters",
  verbose = FALSE
)
```

For each cluster we have a list of potential cell type names and the strength of
their enrichment. We can explore these for each cluster in detail, but as an
initial check let's assign each cluster to the top celltype term:

```{r}
new_ids <- sapply(enrichments, function(df) df$pathway[1])
new_ids[is.na(new_ids)] <- "unknown"
Idents(pbmc) <- "seurat_clusters"
pbmc <- RenameIdents(pbmc, new_ids)
```

```{r fig.height=8, fig.width=15}
plot5 <- DimPlot(pbmc, label = TRUE, repel = TRUE) + NoLegend() + ggtitle("CL enrichment")

plot2 + ggtitle("Label transfer") | plot5
```

As we can see, many of the cell type predictions for our REMO clusters match the
label transfer results. However, some do not. It's important to inspect these
results and combine with other sources of information (gene activity of cell
type marker genes, for example) to decide on a final cell type label assignment.

## Plotting genomic regions

We can plot the frequency of Tn5 integration across regions of the genome for
cells grouped by cluster, cell type, or any other metadata stored in the object
for any genomic region using the `CoveragePlot()` function.
These represent pseudo-bulk accessibility tracks, where signal from
all cells within a group have been averaged together to visualize the DNA 
accessibility in a region (thanks to Andrew Hill for giving the inspiration for
this function in his excellent [blog post](http://andrewjohnhill.com/blog/2019/04/12/streamlining-scatac-seq-visualization-and-analysis/)). Alongside these accessibility tracks we can visualize other
important information including gene annotation, peak coordinates, and genomic
links (if they're stored in the object). See the
[visualization vignette](visualization.html) for more information.

For plotting purposes, it's nice to have related cell types grouped together. We
can automatically sort the plotting order according to similarities across the 
annotated cell types by running the `SortIdents()` function:    

```{r message=FALSE, warning=FALSE}
Idents(pbmc) <- "predicted.id"
pbmc <- SortIdents(pbmc)
```

We can then visualize the DA peaks open in CD4 naive cells and CD14 monocytes, 
near some key marker genes for these cell types, CD4 and LYZ respectively.
Here we'll highlight the DA peaks regions in grey.

```{r message=FALSE, warning=FALSE, out.width='90%', fig.height=6}
CoveragePlot(
  object = pbmc,
  region = "CD4",
  ranges = REMO.v1.GRCh38,
  ranges.group.by = "REMO",
  extend.upstream = 1000,
  extend.downstream = 1000
)
```

```{r message=FALSE, warning=FALSE, out.width='90%', fig.height=6}
CoveragePlot(
  object = pbmc,
  region = "LYZ",
  ranges = REMO.v1.GRCh38,
  ranges.group.by = "REMO",
  extend.upstream = 1000,
  extend.downstream = 5000
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qs2::qs_save(object = pbmc, file = "pbmc_remo.qs2")
```

<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
