---
title: "Calling peaks"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 16 * 1024^3)
options(bitmapType = "cairo")
knitr::opts_knit$set(root.dir = "../../vignette_data/pbmc_vignette")
```

In this tutorial, we demonstrate how to call peaks on a single-cell ATAC-seq
dataset using [MACS3](https://github.com/macs3-project/MACS).

To use the peak calling functionality in Signac you will first need to install
MACS3. This can be done using [pip](https://pypi.org/project/MACS3/) or
[conda](https://anaconda.org/channels/bioconda/packages/macs3/overview), or by building 
the package from [source](https://github.com/macs3-project/MACS). 

MACS3 uses a dedicated parser for single-cell ATAC-seq fragment files, which contains 
the barcode information and the counts of the fragments aligned to the same location 
with the same barcode.

In this demonstration we use scATAC-seq data for human PBMCs. See our
[vignette](pbmc_vignette.html) for the code used to generate this object,
and links to the raw data. First, load the required packages and the
pre-computed Seurat object:

```{r message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)

pbmc <- qs2::qs_read("pbmc.qs2")
DimPlot(pbmc)
```

### Peak calling with `callpeak` mode

Peak calling can be performed using the `CallPeaks()` function on a `Seurat` 
object, `ChromatinAssay5` object, `Fragment2` object, or by giving the path to
a fragment file on disk.

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(pbmc)
head(peaks)
```

### Parallelization across cell groups with `future`

We have observed that the runtime for MACS3 may be longer than previous versions
of MACS that reads in the input as a standard BED file. When calling peaks
across multiple groups of cells, we recommend enabling parallelization in R
using the [future](https://cran.r-project.org/package=future) package
(see `future::plan()` for details on how to configure parallelization).
This will enable peaks to be called in parallel for each group of cells.

```{r message=FALSE, warning=FALSE}
library(future)

# set up workers
plan(multicore, workers = 6)
nbrOfWorkers()
```

To call peaks on each annotated cell type, we can use the `group.by` argument:

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(pbmc, group.by = "predicted.id")
```

The results are returned as a `GRanges` object, with an additional metadata column
listing the cell types that each peak was identified in:

```{r echo=FALSE}
knitr::kable(head(as.data.frame(peaks)))
```

To call peaks for specific group(s) of cells, use the `idents` argument:

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(pbmc, group.by = "predicted.id", idents = "CD16+ Monocytes")
```

By default, `CallPeaks()` will combine all peak calls into one `GRanges` object.
Set `combine.peaks = FALSE` to return a list with a `GRanges` object for each 
group:

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks_separate <- CallPeaks(
  object = pbmc,
  group.by = "predicted.id",
  idents = c("CD4 Naive", "CD4 Memory"),
  combine.peaks = FALSE
)
```

```{r echo=FALSE}
knitr::kable(head(as.data.frame(peaks_separate[[1]])))
knitr::kable(head(as.data.frame(peaks_separate[[2]])))
```

To quantify counts in each peak, you can use the `FeatureMatrix()` function.

We can visualize the cell-type-specific MACS3 peak calls alongside the 10x Cellranger
peak calls (currently being used in the `pbmc` object) with the `CoveragePlot()`
function:

```{r fig.height=6, fig.width=10}
CoveragePlot(
  object = pbmc,
  region = "CD8A",
  ranges = peaks,
  ranges.title = "MACS3"
)
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
