---
title: "Calling peaks"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 16 * 1024^3)
options(bitmapType = 'cairo')
knitr::opts_knit$set(root.dir = "../vignette_data/pbmc_vignette")
```

In this tutorial, we demonstrate how to call peaks on a single-cell ATAC-seq
dataset using [MACS3](https://github.com/macs3-project/MACS).

To use the peak calling functionality in Signac you will first need to install
MACS3. This can be done using [pip](https://pypi.org/project/MACS3/) or
[conda](https://anaconda.org/channels/bioconda/packages/macs3/overview), or by building 
the package from [source](https://github.com/macs3-project/MACS). 

MACS3 uses a dedicated parser for single-cell ATAC-seq fragment files, which contains 
the barcode information and the counts of the fragments aligned to the same location 
with the same barcode. Due to this extra information in the input file, the runtime 
for MACS3 may be longer than MACS2 (old CallPeaks version) that reads in the input
as a standard BED file. Further, it is recommended to enable parallelization in R
using the future package (see the parallel computing [vignette](https://stuartlab.org/signac/articles/future) 
for more details), especially if calling peaks for multiple groups. 

```{r message=FALSE, warning=FALSE}
library(future)

# set up workers
plan(multisession, workers = 6)
nbrOfWorkers()
```

In this demonstration we use scATAC-seq data for human PBMCs. See our
[vignette](pbmc_vignette.html) for the code used to generate this object,
and links to the raw data. First, load the required packages and the
pre-computed Seurat object:


```{r message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)

getwd()

pbmc <- qs2::qs_read("pbmc.qs2")
DimPlot(pbmc)
```

Peak calling can be performed using the `CallPeaks()` function on a `Seurat` 
object, `ChromatinAssay5` object, `Fragment2` object, or a path to fragment file. 
For Seurat objects, peak calling can either be done separately for different 
groups of cells, or performed using data from all the cells. To call peaks 
on each annotated cell type, we can use the `group.by` argument:

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(
    object = pbmc,
    group.by = "predicted.id"
)
```

The results are returned as a `GRanges` object, with an additional metadata column
listing the cell types that each peak was identified in:

```{r echo=FALSE}
knitr::kable(head(as.data.frame(peaks)))
```

If not specified, using the `group.by` argument will call peaks for all unique groups.
To call peaks for specific group(s), use the `idents` arguments.

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks <- CallPeaks(
    object = pbmc,
    group.by = "predicted.id",
    idents = "CD16+ Monocytes"
)
```

By default, `CallPeaks()` will combine all peak calls into 1 `GRanges` object, use the
`combine.peaks = FALSE` argument to return a list with 1 `GRanges` object for each 
group.

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks_separate <- CallPeaks(
    object = pbmc,
    group.by = "predicted.id",
    idents = c("CD4 Naive", "CD4 Memory"), 
    combine.peaks = FALSE
)
```

```{r echo=FALSE}
knitr::kable(head(as.data.frame(peaks_separate)))
```

For MACS version 3.0.4 or above, it is possible to call peaks using the hidden
markov model ([HMMRATAC](https://academic.oup.com/nar/article/47/16/e91/5519166)) mode. 

Do note that this mode requires more computing time and may need additional arguments, see
the MACS3 hmmratac mode (documentation)[https://deepwiki.com/macs3-project/MACS/3.2-atac-seq-analysis-with-hmmratac] 
for more details. Parallelization is recommended if using this mode to call peaks for multiple cell groups.

```{r message=FALSE, warning=FALSE, cache=TRUE}
peaks_hmmratac <- CallPeaks(
    object = pbmc,
    mode = 'hmmratac',
    group.by = 'predicted.id',
    idents = 'CD16+ Monocytes'
)
```

To quantify counts in each peak, you can use the `FeatureMatrix()` function.

We can visualize the cell-type-specific MACS3 peak calls alongside the 10x Cellranger
peak calls (currently being used in the `pbmc` object) with the `CoveragePlot()`
function. Here the Cellranger peaks are shown in grey and the MACS2 peaks in red:

```{r fig.height=6, fig.width=10}
CoveragePlot(
  object = pbmc,
  region = "CD8A",
  ranges = c(peaks, peaks_hmmratac),
  ranges.title = c("MACS3 callpeak", "MACS3 hmmratac")
)
```
<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
