---
title: "Joint RNA and ATAC analysis: SNARE-seq"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

In this vignette we will analyse a single-cell co-assay dataset measuring gene
expression and DNA accessibility in the same cells. This vignette is similar to
the [PBMC multiomic vignette](pbmc_multiomic.html), but demonstrates a similar
joint analysis in a different species and with data gathered using a different technology.

This dataset was published
by [Chen, Lake, and Zhang (2019)](https://www.nature.com/articles/s41587-019-0290-0)
and uses a technology called SNARE-seq. We will look at a dataset from the adult
mouse brain, and the raw data can be downloaded from NCBI GEO here:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE126074 

As the fragment files for this dataset are not publicly available we have
re-mapped the raw data to the mm10 genome and created a fragment file using
[Sinto](https://timoast.github.io/sinto/).

The fragment file can be downloaded
here: https://www.dropbox.com/s/se3uag0cvld4xqg/fragments.sort.bed.gz

The fragment file index can be downloaded here: https://www.dropbox.com/s/kvb8fiewkjz3kzf/fragments.sort.bed.gz.tbi

Code used to create the fragment file from raw data is available here:
https://github.com/timoast/SNARE-seq 

## Data loading

First we create a Seurat object containing two different assays, one containing
the gene expression data and one containing the DNA accessibility data.

To load the count data, we can use the `Read10X()` function from Seurat by first
placing the `barcodes.tsv.gz`, `matrix.mtx.gz`, and `features.tsv.gz` files into
a separate folder.

```{r include=FALSE}
if (!requireNamespace("EnsDb.Mmusculus.v79", quietly = TRUE))
    BiocManager::install("EnsDb.Mmusculus.v79")
```

```{r message=FALSE, warning=FALSE}
library(Signac)
library(Seurat)
library(ggplot2)
library(EnsDb.Mmusculus.v79)
set.seed(1234)

# load processed data matrices for each assay
rna <- Read10X("../vignette_data/snare-seq/GSE126074_AdBrainCortex_rna/", gene.column = 1)
atac <- Read10X("../vignette_data/snare-seq/GSE126074_AdBrainCortex_atac/", gene.column = 1)
fragments <- "../vignette_data/snare-seq/fragments.sort.bed.gz"

# create a Seurat object and add the assays
snare <- CreateSeuratObject(counts = rna)
snare[['ATAC']] <- CreateChromatinAssay(
  counts = atac,
  sep = c(":", "-"),
  genome = "mm10",
  fragments = fragments
)

# extract gene annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)

# change to UCSC style since the data was mapped to hg19
seqlevelsStyle(annotations) <- 'UCSC'
genome(annotations) <- "mm10"

# add the gene information to the object
Annotation(snare[["ATAC"]]) <- annotations
```

## Quality control

```{r message=FALSE, warning=FALSE}
DefaultAssay(snare) <- "ATAC"
snare <- TSSEnrichment(snare)
snare <- NucleosomeSignal(snare)
snare$blacklist_fraction <- FractionCountsInRegion(
  object = snare,
  assay = 'ATAC',
  regions = blacklist_mm10
)
```

```{r fig.width=18, message=FALSE, warning=FALSE}
Idents(snare) <- "all"  # group all cells together, rather than by replicate
VlnPlot(
  snare,
  features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment",
               "nucleosome_signal", "blacklist_fraction"),
  pt.size = 0.1,
  ncol = 5
)
```

```{r}
snare <- subset(
  x = snare,
  subset = blacklist_fraction < 0.03 &
    TSS.enrichment < 20 &
    nCount_RNA > 800 &
    nCount_ATAC > 500
)
snare
```

## Gene expression data processing

Process gene expression data using Seurat

```{r message=FALSE, warning=FALSE}
DefaultAssay(snare) <- "RNA"

snare <- FindVariableFeatures(snare, nfeatures = 3000)
snare <- NormalizeData(snare)
snare <- ScaleData(snare)
snare <- RunPCA(snare, npcs = 30)
snare <- RunUMAP(snare, dims = 1:30, reduction.name = "umap.rna")
snare <- FindNeighbors(snare, dims = 1:30)
snare <- FindClusters(snare, resolution = 0.5, algorithm = 3)
p1 <- DimPlot(snare, label = TRUE) + NoLegend() + ggtitle("RNA UMAP")
```

## DNA accessibility data processing

Process the DNA accessibility data using Signac

```{r message=FALSE, warning=FALSE}
DefaultAssay(snare) <- 'ATAC'

snare <- FindTopFeatures(snare, min.cutoff = 10)
snare <- RunTFIDF(snare)
snare <- RunSVD(snare)
snare <- RunUMAP(snare, reduction = 'lsi', dims = 2:30, reduction.name = 'umap.atac')
p2 <- DimPlot(snare, reduction = 'umap.atac', label = TRUE) + NoLegend() + ggtitle("ATAC UMAP")
```

```{r fig.width=12, message=FALSE, warning=FALSE}
p1 + p2
```

## Integration with scRNA-seq data

Next we can annotate cell types in the dataset by transferring labels from an
existing scRNA-seq dataset for the adult mouse brain, produced by the Allen
Institute.

```{r message=FALSE, warning=FALSE}
# label transfer from Allen brain
allen <- readRDS("../vignette_data/allen_brain.rds")

# use the RNA assay in the SNARE-seq data for integration with scRNA-seq
DefaultAssay(snare) <- 'RNA'

transfer.anchors <- FindTransferAnchors(
  reference = allen,
  query = snare,
  dims = 1:30,
  reduction = 'cca'
)

predicted.labels <- TransferData(
  anchorset = transfer.anchors,
  refdata = allen$subclass,
  weight.reduction = snare[['pca']],
  dims = 1:30
)

snare <- AddMetaData(object = snare, metadata = predicted.labels)
```

```{r}
# label clusters based on predicted ID
new.cluster.ids <- c(
  "L2/3 IT",
  "L4",
  "L6 IT",
  "L5 CT",
  "L4",
  "L5 PT",
  "Pvalb",
  "Sst",
  "Astro",
  "Oligo",
  "Vip/Lamp5",
  "L6 IT.2",
  "L6b",
  "NP"
)
names(x = new.cluster.ids) <- levels(x = snare)
snare <- RenameIdents(object = snare, new.cluster.ids)
snare$celltype <- Idents(snare)
DimPlot(snare, group.by = 'celltype', label = TRUE, reduction = 'umap.rna')
```

## Jointly visualizing gene expression and DNA accessibility

We can visualize both the gene expression and DNA accessibility information at
the same time using the `CoveragePlot()` function. This makes it easy to compare
DNA accessibility in a given region for different cell types and overlay gene 
expression information for different genes.

```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
DefaultAssay(snare) <- "ATAC"
CoveragePlot(snare, region = "chr2-22620000-22660000", features = "Gad2")
```

```{r include=FALSE}
saveRDS(object = snare, file = "../vignette_data/snare.rds")
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
